# March 31st

::: {.callout-tip}
## By the end of today's seminar, you should be able to
 + Discuss the role of Sums of Squares (SS) in variance decomposition
 + "Manually" calculate within, between, and total SS
 + Discuss the issues associated with multiple comparison techniques and the standard adjustment techniques employed to combat the issues
:::


## Rough timeline

 + **09:05–09:55 am** Modeller vs Designer (@sec-mods)
   - Group activity (@sec-ss)
   
 + **09:55–11:05 am** Break
 + **11:00–11:45 am** Multiple comparisons (@sec-multi)
   - Class discussion 
 
 + **11.45–11.55 am** Q&A and peer-share (@sec-peershare07)

   
### Randomly allocated groups

**It's your responsibility to get into these, just look for the cows!**

```{r}
group_allocation(names, seed = 1631, 10)
```


## Modeller vs Designer {#sec-mods}

<p align="center">
<iframe src="slides/05-01_modller-vs-designer.pdf" height="400" width="700" title="Modeller-vs-Designer slides"></iframe>
</p>
<p align="center">
[**CANVAS slides**](https://canvas.auckland.ac.nz/courses/120054/files/15042516?module_item_id=2582752)
</p>


### Partitioning Variation

![📷 [source](Adobe Stock)](runsheets/img/well.jpg)

[**Partitioning Variation**](https://canvas.auckland.ac.nz/courses/120054/files/15015121?module_item_id=2581434) some examples.






### Sums of Squares activity {#sec-ss}

:::{.callout-warning icon=false}
## <i class="fa fa-question-circle" aria-hidden="true"></i> <i class="fa fa-users" aria-hidden="true"></i> <i class="fa fa-pencil-square" aria-hidden="true"></i> Head to this [Google Doc](https://docs.google.com/document/d/1Hby1zVqv39pZSaXwBifZ7W_Mz9fmTWYP8LQ6U8BUQ54/edit?usp=sharing)...

Together, work through the code snippets from [this section of the course guide Sums of squares (SS)](https://biosci738.github.io/BIOSCI738/modelling-experimental-data.html#sums-of-squares-ss), you may have to refer to previous sections (e.g., the [One-Way Analysis of Variance (ANOVA) section from Module 2](https://biosci738.github.io/BIOSCI738/parametric-hypothesis-testing.html#one-way-analysis-of-variance-anova)) if you are unfamiliar with some of the terminology, and discuss/answer the following

 + What are sums of squares (SS)?
 + What role do SS play in statistics?
 + What is meant by the term sequential Sums of Squares? And what effect, if any, might this have on inference?
 + Explain the reasoning/answers above for an [r/ExplainLikeImFive](https://www.reddit.com/r/explainlikeimfive/?rdt=41742) audience.

You may find the pseudo code below useful.

```{r, eval = FALSE}
## template only, you will have to rename variables accordingly
## run segments/lines one at a time to see the steps
data %>% 
    mutate(grand_mean = mean(response)) %>%
   mutate(sstotal = (response - grand_mean) ^2 ) %>%
    group_by(treatment) %>%
   mutate(treatment_mean = mean(response),
          sse = (response - treatment_mean)^2,
          tmp = (treatment_mean - grand_mean)^2) %>%
   summarise(ss_tot = sum(sstotal), ss_e = sum(sse), ss_treat = sum(tmp)) %>%
   pull(ss_treat) %>% sum()
```

:::

::: {.callout-note}
## Bonus activity 
If you have some spare time sketch a concept map or illustration about sequential Sums of Squares.
:::


## Multiple comparisons {#sec-multi}


**Slides**

<p align="center">
<iframe src="slides/05-02_multiple-comparisons.pdf" height="400" width="700" title="Modeller-vs-Designer slides"></iframe>
</p>
<p align="center">
[**CANVAS slides**](https://canvas.auckland.ac.nz/courses/120054/files/15042517?module_item_id=2582753)
</p>


## Q&A and peer-share {#sec-peershare07}

:::{.callout-warning icon=false}
## <i class="fa fa-question-circle" aria-hidden="true"></i> <i class="fa fa-users" aria-hidden="true"></i> <i class="fa fa-pencil-square" aria-hidden="true"></i> Peer-Share
Go over an activity from the class again. Feel free to ask Charlotte a question or two about it!
:::
