# May 2nd

## Rough timeline

### **02:05 pm - 03:55 pm** 

 + **02:05–02.50 pm** Recap (@sec-recap02)
 + **02:50–03:00 pm** Break
 + **03:00–03:55 pm** Introduction to MLE (@sec-mle)
    - Class discussion
    - Group activity

## Recap {#sec-recap02}


:::{.callout-warning icon=false}
## <i class="fa fa-question-circle" aria-hidden="true"></i> <i class="fa fa-users" aria-hidden="true"></i> <i class="fa fa-pencil-square" aria-hidden="true"></i> Revisit...

...the following case studies. In each case, write out your chosen model and create a plot/calculate a measure that demonstrates the "fit" of that model.

 1. **Fry Time** @sec-fry (data from @fries)
 2. **Monkeys** case study 04 from @sec-casestudies (data from @monkey)
 3. **Bird Abundance** third case study from @sec-glmfit (data from @birds)
 
**Be prepared to present your activity to me/the class** (all group members should expect to contribute).

:::


## Maximum Likelihood Estimation (MLE) {#sec-mle}


<p align="center">
<iframe src="slides/mle_summary.html" height="600" width="800" title="Maximum Likelihood Estimation"></iframe>
</p>



:::{.callout-warning icon=false}
## <i class="fa fa-question-circle" aria-hidden="true"></i> <i class="fa fa-users" aria-hidden="true"></i> <i class="fa fa-pencil-square" aria-hidden="true"></i> In your groups...
"Read" and summarise the following cheatsheets.

 1. [MLE (p)](https://github.com/BIOSCI738/cowstats/blob/main/img/mle_p.png?raw=true)
 2. [MLE ($\mu$ & $\sigma$)](https://github.com/BIOSCI738/cowstats/blob/main/img/mle_mu_sigma.png?raw=true)

:::

**Slides**

<p align="center">
<iframe src="slides/07-02_maximum-likelihood.pdf" height="400" width="700" title="Maximum Likelihood Estimation"></iframe>
</p>
<p align="center">
[**CANVAS slides**](https://canvas.auckland.ac.nz/courses/120054/files/15211515?module_item_id=2598102)
</p>


### Case studies

::: {.callout-note}
## We'll work through this example together! Note that this aligns with the content in [this section of the courseguide](https://biosci738.github.io/BIOSCI738/maximum-likelihood-estimation.html).  
:::

```{r, eval =FALSE}
require(tidyverse)
require(patchwork)
#############
## Binomial #
#############

likelihood <- function(theta) dbinom(x = 79, size = 159, prob = theta)
optimise(likelihood, c(0,1), maximum = TRUE)

log_likelihood <- function(theta) dbinom(x = 79, size = 159, prob = theta, log = TRUE)
optimise(log_likelihood, c(0,1), maximum = TRUE)


#############
## Poisson ##
#############

likelihood <- function(lambda) dpois(x = 54, lambda)
optimise(likelihood, c(0,100), maximum = TRUE)

log_likelihood <- function(lambda) dpois(x = 54, lambda, log = TRUE)
optimise(log_likelihood, c(0,100), maximum = TRUE)

################################
## plots for Poisson example ###
################################
lam <- seq(0, 100, length.out = 500)
p1 <- data.frame(lam = lam, likelihood =  dpois(x = 54, lam)) %>%
  ggplot() +
  geom_line(aes(x = lam, y = likelihood)) +
  theme_linedraw() + xlab(expression(lambda)) +
  ylab(expression(paste("L(", ~lambda, ";s)", sep = " "))) +
  geom_vline(xintercept = 54,  col = "#782c26", linewidth = 1.5) +
  ggtitle("Likelihood for X ~ Poisson(54)") +
  annotate(geom = "text", x = 75, y = 0.04, label = "Maximum likelihood \nestimate", col = "#782c26") +
  geom_hline(yintercept = dpois(x = 54, 54),
             lty = 2, col = "#26733a", linewidth = 1.5 ) +
  annotate(geom = "text", x = 5, y = dpois(x = 54, 54) - 0.005 , 
           label = expression(paste(frac("d L","dx"), " = 0")), col = "#26733a")
p2 <- data.frame(lam = lam, loglikelihood =  dpois(x = 54, lam, log = TRUE)) %>%
  ggplot() +
  geom_line(aes(x = lam, y = loglikelihood)) +
  theme_linedraw() + xlab(expression(lambda)) +
  ylab(expression(paste("log(L(", ~lambda, ";s))", sep = " "))) +
  geom_vline(xintercept = 54,  col = "#782c26", linewidth = 1.5) +
  ggtitle("log-likelihood for X ~ Poisson(54)") +
  annotate(geom = "text", x = 70, y = -100, label = "Maximum likelihood \nestimate", col = "#782c26") +
  geom_hline(yintercept = dpois(x = 54, 54, log = TRUE),
             lty = 2, col = "#26733a", linewidth = 1.5 ) +
  annotate(geom = "text", x = 5, y = -20, 
           label = expression(paste(frac("d log(L)","dx"), " = 0")), col = "#26733a")

patchwork::wrap_plots(p1, p2)
```


:::{.callout-warning icon=false}
## <i class="fa fa-question-circle" aria-hidden="true"></i> <i class="fa fa-users" aria-hidden="true"></i> <i class="fa fa-pencil-square" aria-hidden="true"></i> In your groups work through the examples below (both from @sec-glmfit) and discuss the prompts and plots etc.
:::


**Bat Abundance** (subset of data from @bats)

```{r, eval = FALSE}
library(tidyverse)
data <- read_csv("https://raw.githubusercontent.com/STATS-UOA/databunker/master/data/bats.csv")
data %>%
    summarise(mean = mean(TotalCount))
## Assume Poisson distribution
## mle for lambda (across all species) is the mean
#############################################
## So what is the MLE for lambda when x = 100?
## Does this make sense?
## Recall that the (log)likelihood is a function of the parameter(s)
## for some fixed data; it gives the probability of a fixed (set of)
## observation(s) for every possible value of the parameter(s)
data.frame(lambda =  seq(10, 200,by = 1)) %>%
    mutate(ll = dpois(x = 100, lambda , log = TRUE)) %>%
    ggplot(., aes( x = lambda, y = ll)) +
    geom_point() +
    theme_linedraw() + xlab(expression(lambda)) + ylab("log-likelihood value") +
    ggtitle(expression(paste("Log-likelihood for X ~ Poisson(", lambda,"), where x = 100")))
## Now for our data
## log likelihood for sum of poisson random variables
## where observations are the observed number of roosts
## for some unknown parameter lambda
#' @param lambda unknown parameter lambda 
#' @param obvs vector of observations
log_lik_poisson <- function(lambda, obvs){
  llh <- sum(dpois(obvs, lambda, log = TRUE))
  return(llh)
}
## optimise the function above to find the MLE of  unknown param lambda
lam_hat <- optimise(log_lik_poisson, c(10,200) , obvs = data$TotalCount, maximum = TRUE)
lam_hat
## what is out maximum likelihood estimate here?
## plot the log-likelihood for the sum of our observed data
data.frame(lambda =  seq(10, 200, by = 0.5) ) %>%
    mutate(ll = sapply(lambda, log_lik_poisson, obvs = data$TotalCount)) %>%
    ggplot(., aes( x = lambda, y = ll)) +
    geom_point() +
    theme_linedraw() + xlab(expression(lambda)) + ylab("log-likelihood value") +
    geom_vline(xintercept = lam_hat$maximum, size = 2, col = "purple") +
    annotate(geom = "text", lam_hat$maximum, -40000, col = "purple", label = "MLE") +
    ggtitle(expression(paste("Log-likelihood for X ~ Poisson(", lambda,")")))

```

**Lobster Survival** (data from @lobster)

```{r, eval = FALSE}
data <- read_csv("https://raw.githubusercontent.com/STATS-UOA/databunker/master/data/lobster.csv")
data <- data %>%
    mutate(survival = ifelse(survived == 0, "consumed", "alive"))
table(data$survival)
n_survived <- 79
n_total <- 159
## define the likelihood function
## note you will need to fill in the correct values
likelihood <- function(theta) dbinom(x = n_survived, size = n_total, prob = theta)
optimise(likelihood, interval = c(0, 1), maximum = TRUE) ## what have we estimated here?
## define the likelihood function
## note you will need to fill in the correct values
log_likelihood <- function(theta) dbinom(x = n_survived, size = n_total, prob = theta, log = TRUE)
optimise(log_likelihood, interval = c(0, 1), maximum = TRUE)  ## what have we estimated here?
## plot likelihood vs unknown param
p <- seq(0, 1, length.out = 100)
lik <- dbinom(x = n_survived, size = n_total, prob = p)
plot(p, lik, type = "l")
points(p, lik)
## Optimise for parameter p
llh <- function(theta, data){
    sum(dbinom(x = data, size = 1, prob = theta, log = TRUE))
}
optimise(llh, c(0,1), data = data$survived,  maximum = TRUE)
## Now for prob or survival as a function of claw size.
## What is the function I've specified? why the negative here??
llh <- function(par, data){
    -sum(dbinom(x = data$survived,
               size = 1,
               prob = exp(par[1] + par[2]*data$size)/(1 + exp(par[1] + par[2]*data$size)),
               log = TRUE))
}
## find the maximum likelihood estimate for the unknown parameters
optim(fn = llh, par = c(-1, 0.5), data = data)
## Compare to the ungrouped model in the course guide
glm_mod_ug <- glm(survived ~ size, family = "binomial", data = data)
coef(glm_mod_ug)
## Briefly comment on why, or why not, your parameter estimates are the same.
## Has the same model been fitted in each case?
## Plot your fitted models and compare!
```

