# May 23rd

## Rough timeline

### **02:05 pm - 03:55 pm** 

 + **02:05–02.40 pm** Multivariate data (@sec-vizmult)
  + What we're interested in
  + Visualizing it
 + **02:40–03:30 pm** Casestudies [*including a break*] (@sec-mdscasestudies02)
 + **03:30–03:55 pm** Model based multivariate analysis (@sec-modelbasedmult)
   

## Multivariate data {#sec-vizmult}

> Multivariate analyses refers to a suite of statistical techniques designed to analyse data with more than one variable: either multiple dependent variables, or both multiple dependent and independent variables.

A typical sequence of inference for multivariate data would be **Raw Data** $^?\rightarrow$ **Scale/Standardize Raw Data** $^?\rightarrow$ **Create a Similarity/Dissimilarity Matrix**. From the created matrix you might consider **Cluster Analysis** and/or **Ordination** (a fancy word for summarizing/visualizing/reducing multiple variables in/to a lower-dimensional space).

::: {layout-ncol=2}

```{mermaid}
flowchart TD
    A[Cluster Analysis] --> B1(Hierarchical)
    A --> B2(Non hierarchical)
    B1 --> C1[Dendrograms]
    C1 -- ? --> D1[Single]
    C1 -- ? --> D2[Complete]
    C1 -- ? --> D3[UPGMA]
    C1 -- ? --> D4[Ward]
    B2 -- ? --> E1[Kmeans]
    style A fill:#f9f,stroke:#333,stroke-width:4px
```

```{mermaid}
flowchart TD
    A[Ordination] --> B1(Unconstrained)
    A --> B2(Constrained)
    B1 --> C1[Reduced dimensions]
    C1 -- ? --> D1[PCA]
    C1 -- ? --> D2[CA]
    C1 -- ? --> D3[mMDS]
    C1 -- ? --> D4[PCO]
    B2 --> E1[Not covered here]
    style E1 fill:#808080,stroke:#333,stroke-width:0.5px
    style A fill:#f9f,stroke:#333,stroke-width:4px
```
:::

:::{.callout-warning icon=false}
## <i class="fa fa-question-circle" aria-hidden="true"></i> <i class="fa fa-users" aria-hidden="true"></i> <i class="fa fa-pencil-square" aria-hidden="true"></i> In groups of two or three...

The sequences and flow charts above have the occasional $?$ visible in the interim steps. Discuss the decisions & justifications required/made to traverse along each arrow (i.e., pass the $?$).  Elaborate on the what each stage achieves giving specific examples from the courseguide. D
:::

### Visualizing multivariate data

We'll work through the code below together; the data refer to different Pokémon (seen in case study 3 of @sec-model). How useful do you find each plot?

```{r, eval = FALSE}
require(tidyverse)
require(GGally)
url <- "https://raw.githubusercontent.com/STATS-UOA/databunker/master/data/pokemon.csv"
data <- read_csv(url)
## pairs plots for continuous variables
data %>%
  select(where(is.numeric)) |> pairs()
data %>%
  select(where(is.numeric)) |> ggpairs()
## but what if you had a categorical variable
data %>%
  select(legendary, where(is.numeric)) |> ggpairs(aes(color = legendary))
## two (or more) categorical variables
with(data, table(legendary,
    type2)) |> barplot(legend = TRUE)
## conditional
with(data, prop.table(table(legendary,
    type2), margin = 2)) |> barplot(legend = TRUE)
## alluvial plots
require(alluvial)
data$Freq <- 1
mult <- aggregate(Freq ~ type2 + legendary,
    data = data, FUN = sum)
alluvial(mult, freq = mult$Freq, border = NA, col = ifelse(mult$legendary, "red", "blue"))
## What are some potential interpretation pitfalls
## heatmaps
library(pheatmap)
data %>%
  select(where(is.numeric), -Freq, - generation, -number) |> cor() |>
  pheatmap( cluster_rows = TRUE, cluster_cols = TRUE,
    treeheight_row = 0, treeheight_col = 0)
data %>% select(where(is.numeric), -Freq, -generation, -number) |>
  pheatmap( , cluster_rows = TRUE,
    cluster_cols = TRUE, labels_row = data$name, cex = 0.7)
## look at ?pheatmap
library(corrplot)
data %>% select(where(is.numeric), -Freq, - generation, -number) |> cor() |> 
  corrplot(addrect = 3, order = 'hclust')

```



## Casestudies {#sec-mdscasestudies02}

:::{.callout-warning icon=false}
## <i class="fa fa-question-circle" aria-hidden="true"></i> <i class="fa fa-users" aria-hidden="true"></i> <i class="fa fa-pencil-square" aria-hidden="true"></i> Revisit...

...one of the following case studies. Create an informative visualization that appropriately demonstrates the structure of these multivariate data. You may find some useful examples [here](https://epurdom.github.io/Stat131A/book/visualizing-multivariate-data.html#relationships-between-continous-variables).

 1. **Eigenfaces** from @sec-eigenfaces
 2. **Clusterducks** from @sec-clusterducks
 3. **Bird Abundance** first case study from revisited in @sec-mdscasestudies (data from @birds)
 4. **DNA** second case study from @sec-mdscasestudies (data from @dna)
 
**Be prepared to present your activity to me as I'll be prepared to critique!**

:::

## Model based multivariate analysis {#sec-modelbasedmult}

Note this is a bit beyond the scope of this course, but I include it here for fun! We'll go through each example together. Here, we don’t reduce data to correlation or similarity matrix. Can calculate AIC, plot residuals, etc and can directly relate multiple responses to covariates in single analysis....

> The `mvabund` package provides tools for model-based analysis of multivariate abundance data in ecology (@mvabund)

```{r, eval = FALSE}
require(mvabund)
data(Tasmania)
tasmvabund <- mvabund(Tasmania$copepods)
## fit a model
tas.nb <- manyglm(Tasmania$copepods ~ Tasmania$block*Tasmania$treatment, family = "negative.binomial")
summary(tas.nb)
## predict
predict(tas.nb, type = "response")
## model assumptions
plot(tas.nb)
## multivariate hypothesis
anova(tas.nb, p.uni = "adjusted")

```


> The `clustglm` package implements multivariate methods using mixtures combining correspondence analysis (CA), scaling (like multidimensional scaling, or MDS), and pattern detection techniques to analyze data with both categorical and continuous variables (@clustglm)

```{r, eval = FALSE}
## devtools::install_github("vuw-clustering/clustglm")
library(clustglm)
data(cottontails)
## glm
glm.null <- glm(formula = cbind(nsucc,nfail) ~ 1,
                data = cottontails,
                family = "binomial")
glm.null$deviance
logLik(glm.null) 
## clustglm
clustglm.null <- clustglm(formula = cbind(nsucc,nfail) ~ rep(1,74),
                          data = cottontails,
             	          family = "binomial")
clustglm.null$deviance   
clustglm.null$LL   ## calculated slightly differently 
## two clusters...
cg.2clust <- clustglm(formula = cbind(nsucc,nfail) ~ IDclust,
                       family = "binomial",
         	       data = cottontails,
         	       fact4clust = "ID", nclust = 2,
         	       clustfactnames = "IDclust",
         	       start.control = list(randstarts = 100),
         	       verbose = 1)
summary(cg.2clust)
## three clusters?
cg.3clust <- clustglm(formula = cbind(nsucc,nfail) ~ IDclust,
                       family = "binomial",
         	       data = cottontails,
         	       fact4clust = "ID", nclust = 3,
         	       clustfactnames = "IDclust",
         	       start.control = list(randstarts = 100),
         	       verbose = 1)
summary(cg.3clust)
## choose between them
model.list <- list(cg.null = clustglm.null,
                   cg.2clust = cg.2clust,
                   cg.3clust = cg.3clust)
comparison(model.list)
```
