# Week 02 - Lecture 01

## Rough Agenda

### **4:05 pm - 5:55 pm** 

 + Gorilla Gambit (@sec-gorilla)
 + AI in the BOISCI 738 classroom (@sec-ai01) 
 + Group Code of Conduct (@sec-coc)
     - Classwide discussion (@sec-coc01)
     - Group activity (@sec-coc02)
 + Programming Practice (@sec-prog_practice)
 + Peer-share (@sec-peershare01)

## Getting into Groups {#sec-groups}

Each lecture will involve a a few group-based activities/discussion. It is **your responsibility** to organise yourselves into groups (see below). Feel free to reorganise the tables to suit your needs!

::: {.callout-note}
## In groups of ~4

Designate roles:

 - **Facilitator** Responsible for seeking out opinions from all group members and ensuring that everyone has the opportunity to contribute.  
 - **Notetaker/Coder** Responsible for recording the key points or for implementing the code for the group this does **NOT** mean the sole coder, rather the person responsible for running any suggested code etc.
 - **Reporter** Responsible for reporting back to the class on behalf of the group. Seek consensus from the group about the most important aspects of your discussion to share.  
 - **Timekeeper** Responsible for managing time for the group, making sure you stay on track and spend equal time on each question. 

Determine **who is filling each role** by order of upcoming birthdays:  

- The person with the next birthday → **Notetaker**  
- The next birthday after that → **Reporter**  
- The third birthday → **Timekeeper**  
- Next → **Facilitator**  
:::


## Gorilla Gambit {#sec-gorilla}

In your groups **quickly** determine who is filling what role.

Once you have allocated group roles your **Reporter** should come to me to retrieve the first instruction!

## Let's talk AI - for programming {#sec-ai01}

### Who's heard of *bullshit*? 

> *Recently, there has been considerable interest in large language models: machine learning systems which produce human-like text and dialogue. Applications of these systems have been plagued by persistent inaccuracies in their output; these are often called “AI hallucinations”. We argue that these falsehoods, and the overall activity of large language models, is better understood as bullshit in the sense explored by Frankfurt (On Bullshit, Princeton, 2005): the models are in an important way indifferent to the truth of their outputs* 

This is an excerpt from the abstract of @bullshit.

> *If a student is confused about a concept, they can sit with ChatGPT and it will talk to them for hours about that particular concept.*

> *It is a really great tool to create code but also a really great tool to prevent yourself from learning.*

> *It’s teachers’ responsibility to motivate them and make such a problem that [students] are keen to solve and in a way that they actually would like to learn something and realize that they need these skills also in the future.*

> *...there have always been so many ways of cheating, but I don’t think I’ve ever been aware of such an obvious, cheap, and easy way of cheating. Students can get [an AI tool] to answer any question I can ask them at the moment and therefore I have lost my ability to confidently assess any work that students hand in.*

> *I think we need different kinds of professionals with different understandings of computing. Some need to be very deeply involved with how our programming languages work ... others might only need some kind of overall understanding. They are not programmers by themselves, but they still should understand how software is produced.*

All the above are quotes garnered in @instructor.

### Questions


Below *generative AI* refers to tools that can generate text, code, explanations, or other content in response to prompts (e.g., large language models and AI coding assistants).


:::{.callout-warning icon=false}
## <i class="fa fa-question-circle" aria-hidden="true"></i> <i class="fa fa-users" aria-hidden="true"></i> <i class="fa fa-pencil-square" aria-hidden="true"></i> Quick Poll

Open [StatChat](https://canvas.auckland.ac.nz/courses/141056/external_tools/38296) (*log into your at.aucklanduni account*)


 1. **To what extent do you think your peers use generative AI in ways that instructors would consider inappropriate?** Answer Either: A = None, B = Some, C = A Few, D = Many
 2. **I rely on generative AI as a regular part of my workflow for coursework.** Answer Either: A = Strongly Disagree, B = Disagree, C = Neutral, D = Agree, E = Strongly Agree
 3. **Reusing small portions of code from external sources without acknowledgement is unethical.** Answer Either: A = Strongly Disagree, B = Disagree, C = Neutral, D = Agree, E = Strongly Agree

:::


## Group Code of Conduct {#sec-coc}

::: {.callout-tip}
## A classwide agreed group working Code of Conduct
Upon the completion of this activity I will summarise the main themes/suggestions from @sec-coc02 (*that I deem appropriate*). This will become the class-agreed group working Code of Conduct that you are expected to adhere to during each activity.
:::

### Why I think this is important {#sec-coc01}

As a student of University of Auckland student, you are responsible for understanding and abiding by the requirements of the [Student Charter](https://www.auckland.ac.nz/en/students/forms-policies-and-guidelines/student-policies-and-guidelines/student-charter.html).


In this activity we're going to be talking about my and your expectation when working in a group during this class, see [this section of the course guide for further details](https://biosci738.github.io/BIOSCI738/key-information.html#participation).

A Code of Conduct is not just a strange thing the university make you sign. They are a large part of many professional and research-focused bodies beyond university. The following lists just a few examples of societies or institutes you will likely come across during a biostats career in NZ.

 + [The NZ Ecological Society CoC](https://newzealandecology.org/about/council-members/society-rules)
 + [The NZ Statistical Association CoC](https://www.stats.org.nz/code-of-conduct/)
 + [DoCs Code of Ethical Conduct](https://www.doc.govt.nz/about-us/our-policies-and-plans/code-of-ethical-conduct/)

### Write your own {#sec-coc02}

:::{.callout-warning icon=false}
## <i class="fa fa-question-circle" aria-hidden="true"></i> <i class="fa fa-users" aria-hidden="true"></i> <i class="fa fa-pencil-square" aria-hidden="true"></i> Propose your own code of conduct for group work in this class

As a group discuss your expectations of yourself and each other during group work activities. You may want to think about

  + how to ensure all members feel welcome,
  + your expectations around being an active (*and equal*) contributor, and
  + your expectations around peers' use of technologies and generative AI during activities.
  
**Individually** submit your summaries via [StatChat](https://canvas.auckland.ac.nz/courses/141056/external_tools/38296) 
:::

## Good programming practice {#sec-prog_practice}

::: {.callout-tip}
## Abide by good programming practices to avoid future heartaches!

For a more in-depth and general discussion I recommend reading @goodprogramming.
:::


### IMO {#sec-prog_practice01}

Following [this section of the course guide](https://biosci738.github.io/BIOSCI738/accuracy-and-honesty.html#good-coding-practice) let's talk about what good programming practice looks like in this course.

#### You should always start with a **clean work space**. 

Honestly, I think the default RStudio behaviour of restoring `.RData` files etc. just makes everyone lazy...

#### Devise a workflow

> During this course, very likely in other courses you'll be taking this semester and in your future careers you will have to deal with many different datasets, wrangle "dirty" data and deal with data from different sources (at the very least). The key thing is to ensure that ANY ANALYSIS YOU CARRY OUT is TRANSPARENT and FULLY REPRODICIBLE (either for your peers or future you). This is where setting good foundations and devising a well-thought-out workflow is imperative!


```{r, eval = FALSE, echo = FALSE, message = FALSE}
work <- read_csv("runsheets/workflow.csv")
work <- work |> separate_wider_delim(workflow, delim = ",", names_sep = "", too_few = "align_start") |> c() |> unlist()
knitr::kable(sort(table(work)), caption = "A summary of your responses", col.names = c("Answer", "Freq"))
```

#### It's not all about length

::: {.callout-note}
## Lines of code written is NOT a measure of skill! 

Some people thing that a writing a large number of lines of code demonstrates prowess. It does not. Surely we've all added nonsense to essays to "fill up" the word count! 

On the over hand some people strive for carrying out operations in the fewest number of lines possible. This typically makes their work impossible to follow!

:::

I recommend finding a spot you're comfortable in between the following code snippets.^[The former example is a solution posed on [this StackOverflow post](https://stackoverflow.com/questions/19341554/regular-expression-in-base-r-regex-to-identify-email-address) asking about email address validation. The latter snippet is my attempt at a long winded (and far less stable) approach.] Most importantly keep your style readable & consistent! 

```{r}
grepl("^[[:alnum:]._-]+@[[:alnum:].-]+[:alnum:]+$", c("larry@gmail.com", "larry-sally@sally.com", "larry@sally.larry.com", "test@test.com.", "charlottejones-todd"))
```

or

```{r}
email_addresses <- c("larry@gmail.com", "larry-sally@sally.com", "larry@sally.larry.com", "test@test.com.", "charlottejones-todd")
contain_at <- function(x){
  grep("@", x)
}
idx <- contain_at(email_addresses)
correct_email <- email_addresses[idx]
correct_email
contain_notrailing <- function(x){
  grep("^[:alnum:]+", x)
}
idx01 <- contain_notrailing(correct_email)
final_correct_email <- correct_email[idx01]
final_correct_email
```



#### Keep the *inside* thoughts *inside* 

::: {.callout-note}
## An `R` script (or equivalent) is a roadmap to your work. 

You should present the cleanest most direct route you can!

:::

I recommend the latter approach below (if you were going to pass on your solution that is). It's not that each step shouldn't be carried out. On the contrary, exploring your data via printing and plotting it is very important! But when you have a solution, pare down your script! no need to take everyone on your journey.

```{r, eval = FALSE}
## read in data
data <- readr::read_csv("a_valid_filename")
## printing data
print(data)
View(data) ##view opens up new window
data$variable
## Create new data object
newdata <- data$variable
newdata
## plot data
plot(newdata)
## calculate mean
mean <- mean(newdata)
print(mean)
## round
round(mean(newdata))
```

vs

```{r, eval = FALSE}
data <- readr::read_csv("a_valid_filename")
round(mean(data$variable))
```

#### Telltale signs... 

::: {.callout-note}
## An `R` script (or equivalent) showcases your approch!

There are a few quirks that AI insists on including in `R` scripts; with a bit of knowledge these are unnecessary. Rightly or wrongly as soon as I see these in a script I become suspicious!

 1. `print()` statements e.g., `print(variable)`
 2. `cat()` statements e.g., `cat("There are", variable, "students enrolled.\n")`
 
Anyone know why these are telltale signs?

:::

#### Make it EASY for me to give you marks
  
  + **Think** about your audience!
  + Ensure your code is **reproducible**!
  + **Do not include any irrelevant or overly verbose output**; this makes it difficult to find and therefore award the relevant components!
  + Keep your code **tidy** and your plots **neat** and **professional**. For example, use informative, readable axis labels rather than allowing the default behaviour of printing the R object name.  



## Peer-share {#sec-peershare01}

:::{.callout-warning icon=false}
## <i class="fa fa-question-circle" aria-hidden="true"></i> <i class="fa fa-users" aria-hidden="true"></i> <i class="fa fa-pencil-square" aria-hidden="true"></i> Share your approach to the diagnostic coding task

Below is my solution to the diagnostic coding task. This is not the only solution! In your groups run through my approach and share your approach (especially if you got stuck in a few places)! Sharing and talking things through will make things clearer. Trust me! 😉 - (*Also a great task to start practicing for your interactive Oral Assessments*)
:::

```{r, eval = TRUE, echo = TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
url <- "https://raw.githubusercontent.com/STATS-UOA/databunker/master/data/dicots_proportions.csv"
data <- read_csv(url)
data %>%
    select(starts_with(c("Calluna", "Treat"))) %>%
    ## select only heather & treatment cols
    group_by(`Treat!`) %>%
    ## group by treatment so calcs are done by group
    summarise_all(list(mean = mean, sd = sd)) %>%
    ## cal mean and sd of each group
    pivot_longer(!`Treat!`) %>%
    ## flip the data frame to "long" format
    separate(name, c(NA, "year", "calc")) %>%
    ## extract and separate info from name column
    mutate(year = as.numeric(str_remove(year, "vulgaris"))) %>%
    ## keep numeric year info only
    pivot_wider(names_from = "calc", values_from = "value") %>%
    ## data into wider format based on mean & sd
    mutate(`Treat!` = str_replace(`Treat!`, "HB", "BH")) %>%
    ## change treatment label HB to BH
    mutate(`Treat!` = fct_relevel(`Treat!`,c("C", "B", "H", "BH"))) %>%
    ## relevel treatment to help with legend ordering later on
    ggplot(., aes(x = year, y = mean, group = `Treat!`)) + ## set up plot
    geom_point(aes(pch = `Treat!`)) + geom_line(aes(linetype = `Treat!`)) +
    ## add mean points & lines
    ylab("Percentage cover") + xlab("Year") + ## axis labels
    geom_errorbar(aes(ymin = mean - sd/sqrt(6),
                      ymax = mean + sd/sqrt(6)), width = .05) +
    ## add error-bars note we need the standard error of the mean of the proportion
    scale_x_continuous(breaks = seq(8, 12, 1),
                       labels = seq(2008, 2012, 1),
                       expand = expansion(0.01),
                       limits = c(08, 12)) + ## match x-axis labels and limits
    scale_y_continuous(breaks = seq(0, 1, 0.1), labels = seq(0, 100, 10),
                       expand = expansion(0.01),
                       limits = c(0, 0.7)) + ## match y-axis labels and limits
    theme_classic() + ## closest in-built ggplot theme I could find
    theme(legend.title = element_blank()) ## remove legend title

```


