# Week 08 - Lecture 02

## Rough Agenda

### **2:05 pm - 3:55 pm** 

  + Introduction to Bayesian Statistics (@sec-bayes)
  + Casestudies from a Bayesian viewpoint (@sec-bayesestudies)

::: {.callout-tip}
## By the end of today's seminar, you should be able to
 + Distinguish frequentist and Bayesian inference in practice by fitting equivalent regression models in `R` 
 + Specify and justify prior distributions for model parameters
 + Fit, diagnose, and interpret Bayesian regression models using `rstanarm`

:::

NOTE: here we only briefly discuss Bayesian approaches using examples, for more details see [this section of the courseguide](https://biosci738.github.io/BIOSCI738/introduction-to-bayesian-statistics.html) and for a much more rounded discussion I strongly recommend at least Chapter 1 of @bayesrules. This runsheet only skims the basics!

## Introduction to Bayesian Statistics {#sec-bayes}


:::{.callout-warning icon=false}
## <i class="fa fa-question-circle" aria-hidden="true"></i> <i class="fa fa-users" aria-hidden="true"></i> <i class="fa fa-pencil-square" aria-hidden="true"></i> Are you a secret Bayesian?

When prompted via [StatChat](https://canvas.auckland.ac.nz/courses/141056/external_tools/38296) answer either **a**, **b**, or **c** to the following questions^[taken verbatim from @bayesrules] (*record your answers*)

1. When flipping a fair coin, we say that “the probability of flipping Heads is 0.5.” How do you interpret this probability?
    - a. If I flip this coin over and over, roughly 50% will be Heads.  
    - b. Heads and Tails are equally plausible.  
    - c. Both a and b make sense.  

2. An election is coming up and a pollster claims that candidate A has a 0.9 probability of winning. How do you interpret this probability?
    - a. If we observe the election over and over, candidate A will win roughly 90% of the time.  
    - b. Candidate A is much more likely to win than to lose.  
    - c. The pollster’s calculation is wrong. Candidate A will either win or lose, thus their probability of winning can only be 0 or 1.  

3. Consider two claims.  
    (1) Zuofu claims that he can predict the outcome of a coin flip. To test his claim, you flip a fair coin 10 times and he correctly predicts all 10.  
    (2) Kavya claims that she can distinguish natural and artificial sweeteners. To test her claim, you give her 10 sweetener samples and she correctly identifies each.  
    In light of these experiments, what do you conclude?  
    - a. You’re more confident in Kavya’s claim than Zuofu’s claim.  
    - b. The evidence supporting Zuofu’s claim is just as strong as the evidence supporting Kavya’s claim.  

4. Suppose that during a recent doctor’s visit, you tested positive for a very rare disease. If you only get to ask the doctor one question, which would it be?  
    - a. What’s the chance that I actually have the disease?  
    - b. If in fact I don’t have the disease, what’s the chance that I would’ve gotten this positive test result?


:::

::: {.callout-note}
## The simplified distinction

 + **Frequentist**: Parameter(s) is/are **fixed** but unknown (i.e., there is one *true* value of each parameter). Inference is based on the likelihood function. Estimation (e.g., MLE) chooses the value of the parameter(s) that maximises this likelihood (recall @sec-mle). 

 + **Bayesian**: Parameter(s) is/are treated as a random variable with a probability distribution. Inference combines the likelihood with a prior distribution via [Bayes’ theorem](https://biosci738.github.io/BIOSCI738/introduction-to-bayesian-statistics.html#bayes-rule). The likelihood plays the same role as in the frequentist setting (i.e., it reflects information from the data) but is now updated with prior beliefs about the parameter(s) to produce the posterior distribution (which we use for inference).


**Both approaches are built on the same likelihood function!**

:::


:::{.callout-note collapse=true}
## Quick Check: Frequentist vs Bayesian Thinking
<div id="bayes-freq-quiz">



<form>

<p><strong>1. In a frequentist approach, how is a parameter treated?</strong></p>
<label><input type="radio" name="q1" value="a"> It is a random variable with a distribution</label><br>
<label><input type="radio" name="q1" value="b"> It is fixed but unknown</label><br>
<label><input type="radio" name="q1" value="c"> It changes every time we collect data</label>

<p><strong>2. In a Bayesian approach, how is a parameter treated?</strong></p>
<label><input type="radio" name="q2" value="a"> It is fixed and cannot be described with probabilities</label><br>
<label><input type="radio" name="q2" value="b"> It is only estimated using maximum likelihood</label><br>
<label><input type="radio" name="q2" value="c"> It is a random variable with a probability distribution</label>

<p><strong>3. What role does the likelihood play?</strong></p>
<label><input type="radio" name="q3" value="a"> It is only used in Bayesian methods</label><br>
<label><input type="radio" name="q3" value="b"> It is only used in frequentist methods</label><br>
<label><input type="radio" name="q3" value="c"> It is used in both approaches</label>

<p><strong>4. What is a key feature of Bayesian analysis?</strong></p>
<label><input type="radio" name="q4" value="a"> It ignores the data</label><br>
<label><input type="radio" name="q4" value="b"> It combines prior information with data</label><br>
<label><input type="radio" name="q4" value="c"> It assumes parameters are always zero</label>


<br><br>
<button type="button" onclick="checkAnswers()">Check Answers</button>

</form>

<script>
function checkAnswers() {
  const answers = {
    q1: "b",
    q2: "c",
    q3: "c",
    q4: "b"
  };

  // Reset styles
  document.querySelectorAll("label").forEach(l => {
    l.style.color = "black";
    l.style.fontWeight = "normal";
  });

  for (let i = 1; i <= 6; i++) {
    const selected = document.querySelector('input[name="q' + i + '"]:checked');
    const options = document.querySelectorAll('input[name="q' + i + '"]');

    options.forEach(opt => {
      const label = opt.parentElement;

      // highlight correct answer
      if (opt.value === answers["q" + i]) {
        label.style.fontWeight = "bold";
      }

      // colour selected answer
      if (selected && opt === selected) {
        if (opt.value === answers["q" + i]) {
          label.style.color = "green";
        } else {
          label.style.color = "red";
        }
      }
    });
  }
}
</script>

</div>

:::

### A Bayesian Beta-Binomial model


We'll work through this example together. Note that this aligns with the content in [this section of the courseguide](https://biosci738.github.io/BIOSCI738/introduction-to-bayesian-statistics.html); however, here we go beyond what is covered there!  


Let's take a look again at the **Lobster Survival** data from @lobster. The NULL model equivalent to the ungrouped model we considered in @sec-glmfit was

```{r, eval = TRUE, echo = FALSE, message = FALSE}
library(tidyverse)
data <- read_csv("https://raw.githubusercontent.com/STATS-UOA/databunker/master/data/lobster.csv")
glm_mod_bern <- glm(survived ~ 1, family = "binomial", data = data)
equatiomatic::extract_eq(glm_mod_bern)
```

In other words, let $y$ be either 1 or 0 depending on survival then $y \sim \text{Binomial}(n = 1, p)$ (i.e., $y \sim \text{Bernoulli}(p)$) where $p$ is the probability of survival.

Here,  $y \sim \text{Binomial}(n = 1, p)$ is the **likelihood** (see @sec-mle) and $p$ is an **unknown parameter** we wish to estimate. Under a Bayesian framework we assume that $p$ is a random variable, with it's own distribution!

**So, what do we know about $p$?** It is a probability and therefore bounded between 0 and 1. At this stage we have no further information (i.e., nothing about what value this probability might be).


:::{.callout-warning icon=false}
## <i class="fa fa-question-circle" aria-hidden="true"></i> <i class="fa fa-users" aria-hidden="true"></i> <i class="fa fa-pencil-square" aria-hidden="true"></i> Suggest...

...a candidate distribution for $p$  via [StatChat](https://canvas.auckland.ac.nz/courses/141056/external_tools/38296) and give your reasoning.

This distribution is your **prior** distribution and reflect your beliefs about $p$.

:::

Recall, your **prior** distribution reflects your beliefs about $p$.

```{r, eval = FALSE}
library(bayesrules)
## The data (i.e., likelihood)
plot_binomial_likelihood(y = 79, n = 159)
## What about a Beta prior on p
plot_beta(10,3) ##??
plot_beta(0.5,0.5) ## favours extremes (actually called a Jeffreys prior)
plot_beta(1,1) ##?? (all equally likely)
## The posterior with an uninformative prior
plot_beta_binomial(alpha = 1, beta = 1, y = 79, n = 159)
## What about an "extreme favouring" prior
plot_beta_binomial(alpha = 0.5, beta = 0.5, y = 79, n = 159)
## The data usurps the prior!! As it should
## Inference
summarize_beta_binomial(alpha = 1, beta = 1, y = 79, n = 159)
## The estimate we're interested in is the posterior mean
## How does this compare to our MLE approach?
log_likelihood <- function(theta) dbinom(x = 79, size = 159, prob = theta, log = TRUE)
optimise(log_likelihood, c(0,1), maximum = TRUE)
```

Read more about prior sensitivity (i.e., your choice of prior affecting results) in [this section of the courseguide](https://biosci738.github.io/BIOSCI738/introduction-to-bayesian-statistics.html#prior-sensitivity). 

What about frequentist approach vs Bayesian? 

> Typically, Bayesian and frequentist estimates will always agree if there is sufficient data, so long as the likelihood is not explicitly ruled out by the prior.— Olivier Gimenez (@olivier)

## Casestudies from a Bayesian viewpoint {#sec-bayesestudies}

### Activity: Choosing a prior distribution

In your groups discuss/answer the scenarios below. Be prepared to answer questions as I wander around the class and **make sure your nominated reporter is ready to report to the class when called upon!**


 1. **Test Scores Under a New Teaching Method** A school is testing a new math teaching method. A sample of students is taught using the new method, and their standardized test scores are recorded. Assume test scores are normally distributed with known variance $\sigma^2 = 100$ (i.e., standard deviation = 10). We're interested in the average score $\mu$ under the new method as follows $y_i \sim \mathcal{N}(\mu, \sigma^2)$. What might be a suitable distribution to reflect the prior belief about the average score $\mu$.
 
 2. **Kiwi Call Monitoring** Wildlife biologists are studying nocturnal kiwi activity in a forest reserve. On several nights, researchers record the number of kiwi calls heard per hour from a fixed listening post. Let $y_i$ be the number of kiwi calls heard during hour $i$, modeled as $y_i \sim \text{Poisson}(\lambda)$. What might be a suitable distribution to reflect the prior belief about the average call rate $\lambda$. 

Now, a statistician has come along and said that things will be a lot easier (ask Charlotte why if you're interested) and said that an appropriate prior for the average score (scenario 1) is $\mu \sim \mathcal{N}(\mu_0, \tau^2)$ and for the average call rate (scenario 2) is $\lambda \sim \text{Gamma}(\alpha, \beta)$. In addition, the principal of the school in scenario 1 believes students taught with the new method will average around 75 points, but the level of certainty varies depending on context. And Ecologists familiar with the region in scenario 2 suggest that the rate of kiwi calls typically ranges from 2 to 4 calls per hour. 

Assuming these prior distributions and the extra information above discuss/answer the following
  
  a. Based on the prior knowledge, how might you choose values for the parameters of each prior distribution?
  
  b. What prior reflects strong certainty?
  
  c. What prior reflects strong uncertainty?


*[NOTE: Scenarios were created with the assistance of ChatGPT (OpenAI, 2025).]*




### Casestudy

Together we'll work through the section below (previously seen example from @sec-glmfit) and discuss the prompts and plots etc. 

*NOTE that this is quite advanced and touches on the boundary of what is beyond the scope of this course, but you're up to it!*

#### A Gamma-Poisson Model (i.e., Negative Binomial)

**Bat Abundance** (subset of data from @bats)

Recall the following model from @sec-activbats, which we fitted in a frequentist framework

$$
\log ({ E( \operatorname{TotalCount} ) })  = \alpha + \beta_{1}(\operatorname{SpeciesID}_{\operatorname{Rhinolophus\ hipposideros}})
$$

where $Y = y$ was the nest count and $y \sim \text{NegBin}(\mu, \phi)$, with $\mu$ the expected count ($E( \operatorname{TotalCount} )$) and $\phi$ the dispersion parameter (i.e., how much variance exceeds the mean).

```{r, eval = FALSE}
library(tidyverse)
bats <- read_delim("https://raw.githubusercontent.com/STATS-UOA/databunker/master/data/bats.csv")
## frequentist approach
nbmod <- MASS::glm.nb(TotalCount ~ SpeciesID, data = bats)
nbmod |> summary() |> coef()
```

Just a recap, why did we choose a Negative Binomial model? Because the was a lot more variation than we'd expect under a Poisson model. We saw this in @sec-activbats, but let's use simulation to *show* this again.

```{r, eval = FALSE}
## observed mean
mean(bats$TotalCount) ## ~174
## observed variance
var(bats$TotalCount) ## a LOT higher
## simulate from a Poisson with lambda = observed mean
sim <- replicate(100, rpois(100, mean(bats$TotalCount))) ## quicker than a 'for loop'
## wrangle and plot
sim |> as.data.frame() %>% 
  pivot_longer(., everything(), names_to = "column", values_to = "value") %>%
  ggplot(aes(x = value, group = column)) +
  geom_density(linewidth = 0.2, alpha = 0.1) +
  geom_density(data = bats, aes(x = TotalCount), color = "purple", 
               linewidth = 1, inherit.aes = FALSE) 
  theme_minimal()
## Clearly the observed (purple) doesn't match the assumed Poisson (all the black lines)
```

OK, back to Bayesian modelling. We have $y \sim \text{NegBin}(\mu, \phi)$. We have two unknown parameters $\mu$ and $\phi$. Let's first consider a NULL (i.e., intercept only) model $\log (\mu)  = \alpha$.

We use the following priors

   1. $\alpha \sim \text{N}(0,5)$ and
   2. $\phi \sim \text{Exp}(1)$.
   
Do these seem appropriate?

```{r, eval = FALSE}
plot_normal(0, 5)
plot_gamma(1, 1) ## recall that exponential is a special case of Gamma with shape = 1 
```

```{r, eval = FALSE}
library(rstanarm)
fit_null <- stan_glm(
  TotalCount ~ 1,
  family = neg_binomial_2(), ## Likelihood
  data = bats,
  prior = normal(0, 5),             ## Prior on log(mu) (i.e., alpha)
  prior_aux = exponential(1),       ## Prior on phi (overdispersion)
  chains = 4, iter = 2000, seed = 123
)
## coefficients
print(fit_null, digits = 2)
```


Now, let's consider a model with the factor explanatory variable `SpeciesID`. Here $\log(\mu)  = \alpha + \beta_{1}(\operatorname{SpeciesID}_{\operatorname{Rhinolophus\ hipposideros}})$. So now we have three unknown parameters $\alpha$, $\beta_1$ and $\phi$.

We use the following priors

   1. $\alpha \sim \text{N}(0, 2.5)$, 
   2. $\beta_1 \sim  \text{N}(0, 5)$, and
   3. $\phi \sim \text{Exp}(1)$.
   
Fitting the model:


```{r, eval = FALSE}
fit_species <- stan_glm(
  TotalCount ~ SpeciesID,
  family = neg_binomial_2(),
  data = bats,
  prior = normal(0, 2.5),           # Prior on coefficients (log scale)
  prior_intercept = normal(0, 5),   # Prior on intercept
  prior_aux = exponential(1),       # Prior on phi (overdispersion)
  chains = 4, iter = 2000, seed = 123
)
## coefficients
print(fit_species, digits = 2)
## compare to
nbmod |> summary() |> coef()
```

**Assessing model fit** the full details of this are beyond the scope of this course, but let's use a common sense check of *does our data look similar to the model*. Special Bayesian terminology for this might be a *posterior predictive check*, we can do this in one step using `pp_check()`:

```{r, eval = FALSE}
pp_check(fit_species)
```

::: {.callout-note}
## Bonus activities
Revisit one of the following case studies and fit an appropriate model in a Bayesian framework

 1. The **Monkeys** case study from @sec-casemonkey (data from @monkey)
 2. The **Bird Abundance** case study from @sec-glmfit (data from @birds)
 
 
Fit an appropriate model to your [Significance Article](https://canvas.auckland.ac.nz/courses/141056/assignments/490408) data in a Bayesian framework! Remember, your project could just as easily be focused on communicating statistical methodology to a general audience as it could be based on some analyses ;)

:::



