# Week 02 - Lecture 02

## Rough Agenda

### **2:05 pm - 3:55 pm** 

 + The power of simulation (@sec-simulation)
 + Reverse engineer (@sec-draw)


### Before we begin **sort yourselves into groups** following @sec-groups

::: {.callout-tip}
## Before modelling, you should first be able to simulate!
 
 + Translate a data-generating *story* into a formal (linear) model
 + Identify and justify the assumptions required for simulation
 + Separate assumptions about covariates from assumptions about errors
 + Explore the consequences of assumption violations

:::


## Simulation {#sec-simulation}


:::{.callout-warning icon=false}
## <i class="fa fa-question-circle" aria-hidden="true"></i> <i class="fa fa-users" aria-hidden="true"></i> <i class="fa fa-pencil-square" aria-hidden="true"></i> What is a Model?

Discuss, then open [StatChat](https://canvas.auckland.ac.nz/courses/141056/external_tools/38296) and answer!

:::

### Lets improv!

I'm going to *attempt* to do this on the fly. So, when I prompt you yell out an answer to the questions below!

 + Something measurable we might be interested in (e.g., commute time to university)
 + Something that might affect this *thing* of interest (e.g., distance to university)
 + Something else that might affect this *thing* of interest (e.g., mode of transport)
 + Anything else...
 
 
 
### Simulate your own scenario

In your groups 

 + Define a measurable outcome of interest
 + Identify predictors (something you think will influence to outcome)
 + What do you think this relationship will be (e.g., *linear*), formulate this systematic relationship
 + Now, think about the random variation (error)
 + Simulate your outcome.
 + What are your assumption? Write them down!
 
Be prepared to discuss your progress/thoughts with me as I wander around the class.

Extra things to think about: *Could there be non-linear relationships?*, *Could there be interaction effects?*, *Could some predictors be correlated?*, *Could errors be non-constant or clustered?*
 
:::{.callout-warning icon=false}
## <i class="fa fa-question-circle" aria-hidden="true"></i> <i class="fa fa-users" aria-hidden="true"></i> <i class="fa fa-pencil-square" aria-hidden="true"></i> 

Once you've simulated a scenario you're happy with (**individually**) paste your code and assumptions into [StatChat](https://canvas.auckland.ac.nz/courses/141056/external_tools/38296).

:::

## From numbers to plots {#sec-draw}

Below is some `R` output showing data summaries and linear model summary output. In your groups, read through and interpret the output then, **on the whiteboards** sketch the model fit (including uncertainty), for each case.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
## data source: https://docs.google.com/spreadsheets/d/1YSJ4ypkYLq6j1mIBJCgUHhHjJZQ0Rkfe1qW2WC5HLiw/edit#gid=748627588
library(tidyverse)
data <- read_csv("https://raw.githubusercontent.com/STATS-UOA/databunker/master/data/marvel_movies.csv") %>%
	mutate(critics_score = as.numeric(str_replace_all(`critics % score`, "%", ""))) %>%
	mutate(budget_recovered = as.numeric(str_replace_all(`% budget recovered`, "%", "")))
```


::: {.callout-caution collapse="true"}
## From numbers to plots - Code and Output

```{r}

data
str(data)
##### ##### ##### ###
##### Model 1 #######
##### ##### ##### ###
data %>%
	lm(budget_recovered ~ critics_score, data = .) |>
	summary()
##### ##### ##### ###
##### Model 2 #######
##### ##### ##### ###
names(table(data$category))
mod <- data %>%
	lm(budget_recovered ~  category, data = .) 
mod |> summary()
##### ##### ##### ###
##### Model 3 #######
##### ##### ##### ###
mod <- data %>%
	lm(budget_recovered ~ critics_score + category, data = .) 
mod |> summary()
##### ##### ##### ###
##### Model 4 #######
##### ##### ##### ###
data <- data %>%
	rename(., "worldwide" = `worldwide gross ($m)`, "domestic" = `domestic gross ($m)`,
				 "international" = `international gross ($m)`)
mod <- lm(worldwide ~ 0 + domestic + international, data = data) 
mod |> summary()
##### ##### ##### ###
##### Model 5 #######
##### ##### ##### ###
mod <- lm(worldwide ~ budget*critics_score, data = data) 
mod |> summary()
```

:::





