# Week 08 - Lecture 01

## Rough Agenda

### **04:05 pm - 05:55 pm** 

 + Recap *model fit* (@sec-recap02)
 + Introduction to MLE (@sec-mle)
    - Class discussion
    - Group activity
    
    
::: {.callout-tip}
## By the end of today's seminar, you should be able to
 + Explain the role of maximum likelihood estimation
 + Define a (log) likelihood function for a Poisson and Binomial model
 + Write out an appropriate (log) likelihood function in `R` for a Poisson and Binomial model and optimise accordingly

:::

## Recap *model fit* {#sec-recap02}


 Revisit the following case studies. In each case, write out your chosen model and create a plot/calculate a measure that demonstrates the "fit" of that model.

 1. The **Monkeys** case study from @sec-casemonkey (data from @monkey)
 2. The **Bird Abundance** case study from @sec-glmfit (data from @birds)
 
**Be prepared to present your model & plots to me as I wander around class**!



## Maximum Likelihood Estimation (MLE) {#sec-mle}


**Group Activity** "Read" the following cheatsheets

 1. [MLE (p)](https://github.com/BIOSCI738/cowstats/blob/main/img/mle_p.png?raw=true)
 2. [MLE ($\mu$ & $\sigma$)](https://github.com/BIOSCI738/cowstats/blob/main/img/mle_mu_sigma.png?raw=true)

and summarise each in your own words. **Be prepared to share your thoughts as I wander around class**!

<p align="center">
<iframe src="slides/mle_summary.pdf" height="400" width="700" title="Maximum Likelihood Estimation"></iframe>
</p>

<p align="center">
<iframe src="slides/07-02_maximum-likelihood.pdf" height="400" width="700" title="Maximum Likelihood Estimation"></iframe>
</p>



::: {.callout-note }
## Maximum Likelihood Estimation (MLE) 

- We assume a distribution for the data (e.g., Binomial with unknown parameter $p$, or Normal with unknown parameters $\mu$ & $\sigma$).  

- The **likelihood** is the  measures the *plausibility* of parameter values  given observed data

$$
L(\theta; y) = \prod_{i=1}^n f(\theta, y_i  )
$$

- Products are hard to work with, so we take logs

$$
\ell(\theta; y) = \log L(\theta;y) = \sum_{i=1}^n \log f(\theta, y_i )
$$

- This turns a **product into a sum**, which is:
  - numerically stable  
  - easier to compute and optimise  

- Maximising likelihood is the same as maximising log-likelihood.

- In practice, we often **minimise the negative log-likelihood** ($-\ell(\theta; y)$)


- So the workflow is:
  1. Write down likelihood (product)  
  2. Take logs $\rightarrow$ sum  
  3. Multiply by $-1$  
  4. Use `optim()` to **minimise the negative log-likelihood**

---

**Key idea:**
$$
\text{maximise } L(\theta;y) \;\equiv\; \text{maximise } \ell(\theta;y) \;\equiv\; \text{minimise } -\ell(\theta;y)
$$
:::


:::{.callout-warning icon=false}
## <i class="fa fa-question-circle" aria-hidden="true"></i> <i class="fa fa-users" aria-hidden="true"></i> <i class="fa fa-pencil-square" aria-hidden="true"></i> Quiz

Given values `y <- c(5.2, 4.8, 6.1, 5.5, 5.9)` and assuming a Normal distribution function, what is the

 a. value of the likelihood for parameter values `mu = 5.5` and `sd = 0.5`,
 b. value of the log-likelihood for parameter values `mu = 5.5` and `sd = 0.5`,
 c. value of the negative log-likelihood for parameter values `mu = 5.5` and `sd = 0.5`,
 d. the estimated parameter values of $\hat{\mu}$ and $\hat{\sigma}$ (i.e., mean and standard deviation) obtained via `optim()`
 e. maximum likelihood value (i.e., the value of the objective function from d.)
 
Here are some `R` functions you might find useful: `dnorm()`, `prod()`, `sum()`!

```{r, eval = FALSE, echo = FALSE}
dnorm(y, mean = 5.5, sd = 0.5) |> prod()
dnorm(y, mean = 5.5, sd = 0.5, log = TRUE) |> sum()
nll <- function(mu, sigma) {
  -sum(dnorm(y, mean = mu, sd = sigma, log = TRUE))
}
nll(5.5, 0.5)
optim(c(5, 1), function(par) nll(par[1], par[2]))

```


When prompted answer via [StatChat](https://canvas.auckland.ac.nz/courses/141056/external_tools/38296) 
:::


### Case studies

We'll work through this example together! Note that this aligns with the content in [this section of the courseguide](https://biosci738.github.io/BIOSCI738/maximum-likelihood-estimation.html).  

```{r, eval =FALSE}
require(tidyverse)
require(patchwork)
#############
## Binomial #
#############

likelihood <- function(theta) dbinom(x = 79, size = 159, prob = theta)
optimise(likelihood, c(0,1), maximum = TRUE)

log_likelihood <- function(theta) dbinom(x = 79, size = 159, prob = theta, log = TRUE)
optimise(log_likelihood, c(0,1), maximum = TRUE)

###########
## plot ###
###########
p <- seq(0, 1, length.out = 500)
data.frame(p = p, Likelihood =  dbinom(x = 79, size = 159, p)) %>%
  ggplot() +
  geom_line(aes(x = p, y = Likelihood)) +
  theme_linedraw() + xlab(expression(theta)) +
  ylab(expression(paste("L(", ~theta, ";s)", sep = " "))) +
  geom_vline(xintercept = 79/159,  col = "#782c26", size = 1.5) +
  ggtitle("Likelihood for X ~ Binomial(59, 179)") +
  annotate(geom = "text", x = 0.75, y = 0.045, label = "Maximum likelihood estimate", col = "#782c26") +
  geom_hline(yintercept = dbinom(x = 79, size = 159, 0.49),
             lty = 2, col = "#26733a", linewidth = 1.5 ) +
  annotate(geom = "text", x = 0.1, y = 0.05, 
           label = expression(paste(frac("d L","dx"), " = 0")), col = "#26733a")


#############
## Poisson ##
#############

likelihood <- function(lambda) dpois(x = 54, lambda)
optimise(likelihood, c(0,100), maximum = TRUE)

log_likelihood <- function(lambda) dpois(x = 54, lambda, log = TRUE)
optimise(log_likelihood, c(0,100), maximum = TRUE)


################################
## plots for Poisson example ###
################################
lam <- seq(0, 100, length.out = 500)
p1 <- data.frame(lam = lam, likelihood =  dpois(x = 54, lam)) %>%
  ggplot() +
  geom_line(aes(x = lam, y = likelihood)) +
  theme_linedraw() + xlab(expression(lambda)) +
  ylab(expression(paste("L(", ~lambda, ";s)", sep = " "))) +
  geom_vline(xintercept = 54,  col = "#782c26", linewidth = 1.5) +
  ggtitle("Likelihood for X ~ Poisson(54)") +
  annotate(geom = "text", x = 75, y = 0.04, label = "Maximum likelihood \nestimate", col = "#782c26") +
  geom_hline(yintercept = dpois(x = 54, 54),
             lty = 2, col = "#26733a", linewidth = 1.5 ) +
  annotate(geom = "text", x = 5, y = dpois(x = 54, 54) - 0.005 , 
           label = expression(paste(frac("d L","dx"), " = 0")), col = "#26733a")
p2 <- data.frame(lam = lam, loglikelihood =  dpois(x = 54, lam, log = TRUE)) %>%
  ggplot() +
  geom_line(aes(x = lam, y = loglikelihood)) +
  theme_linedraw() + xlab(expression(lambda)) +
  ylab(expression(paste("log(L(", ~lambda, ";s))", sep = " "))) +
  geom_vline(xintercept = 54,  col = "#782c26", linewidth = 1.5) +
  ggtitle("log-likelihood for X ~ Poisson(54)") +
  annotate(geom = "text", x = 70, y = -100, label = "Maximum likelihood \nestimate", col = "#782c26") +
  geom_hline(yintercept = dpois(x = 54, 54, log = TRUE),
             lty = 2, col = "#26733a", linewidth = 1.5 ) +
  annotate(geom = "text", x = 5, y = -20, 
           label = expression(paste(frac("d log(L)","dx"), " = 0")), col = "#26733a")

patchwork::wrap_plots(p1, p2)
```

### Activities {#sec-activbats}

In your groups work through the examples below and discuss the prompts and plots etc. Be prepared to answer questions as I wander around the class.


**Bat Abundance** (subset of data from @bats)

```{r, eval = FALSE}
library(tidyverse)
data <- read_csv("https://raw.githubusercontent.com/STATS-UOA/databunker/master/data/bats.csv")
data %>%
    summarise(mean = mean(TotalCount))
## Assume Poisson distribution
## mle for lambda (across all species) is the mean
#############################################
## So what is the MLE for lambda when x = 100?
## Does this make sense?
## Recall that the (log)likelihood is a function of the parameter(s)
## for some fixed data; it gives the probability of a fixed (set of)
## observation(s) for every possible value of the parameter(s)
data.frame(lambda =  seq(10, 200,by = 1)) %>%
    mutate(ll = dpois(x = 100, lambda , log = TRUE)) %>%
    ggplot(., aes( x = lambda, y = ll)) +
    geom_point() +
    theme_linedraw() + xlab(expression(lambda)) + ylab("log-likelihood value") +
    ggtitle(expression(paste("Log-likelihood for X ~ Poisson(", lambda,"), where x = 100")))
## Now for our data
## log likelihood for sum of poisson random variables
## where observations are the observed number of roosts
## for some unknown parameter lambda
#' @param lambda unknown parameter lambda 
#' @param obvs vector of observations
log_lik_poisson <- function(lambda, obvs){
  llh <- sum(dpois(obvs, lambda, log = TRUE))
  return(llh)
}
## optimise the function above to find the MLE of  unknown param lambda
lam_hat <- optimise(log_lik_poisson, c(10,200) , obvs = data$TotalCount, maximum = TRUE)
lam_hat
## what is out maximum likelihood estimate here?
## plot the log-likelihood for the sum of our observed data
data.frame(lambda =  seq(10, 200, by = 0.5) ) %>%
    mutate(ll = sapply(lambda, log_lik_poisson, obvs = data$TotalCount)) %>%
    ggplot(., aes( x = lambda, y = ll)) +
    geom_point() +
    theme_linedraw() + xlab(expression(lambda)) + ylab("log-likelihood value") +
    geom_vline(xintercept = lam_hat$maximum, size = 2, col = "purple") +
    annotate(geom = "text", lam_hat$maximum, -40000, col = "purple", label = "MLE") +
    ggtitle(expression(paste("Log-likelihood for X ~ Poisson(", lambda,")")))

```

**Lobster Survival** (data from @lobster)

```{r, eval = FALSE}
data <- read_csv("https://raw.githubusercontent.com/STATS-UOA/databunker/master/data/lobster.csv")
data <- data %>%
    mutate(survival = ifelse(survived == 0, "consumed", "alive"))
table(data$survival)
n_survived <- 79
n_total <- 159
## define the likelihood function
## note you will need to fill in the correct values
likelihood <- function(theta) dbinom(x = n_survived, size = n_total, prob = theta)
optimise(likelihood, interval = c(0, 1), maximum = TRUE) ## what have we estimated here?
## define the likelihood function
## note you will need to fill in the correct values
log_likelihood <- function(theta) dbinom(x = n_survived, size = n_total, prob = theta, log = TRUE)
optimise(log_likelihood, interval = c(0, 1), maximum = TRUE)  ## what have we estimated here?
## plot likelihood vs unknown param
p <- seq(0, 1, length.out = 100)
lik <- dbinom(x = n_survived, size = n_total, prob = p)
plot(p, lik, type = "l")
points(p, lik)
## Optimise for parameter p
llh <- function(theta, data){
    sum(dbinom(x = data, size = 1, prob = theta, log = TRUE))
}
optimise(llh, c(0,1), data = data$survived,  maximum = TRUE)
## Now for prob or survival as a function of claw size.
## What is the function I've specified? why the negative here??
llh <- function(par, data){
    -sum(dbinom(x = data$survived,
               size = 1,
               prob = exp(par[1] + par[2]*data$size)/(1 + exp(par[1] + par[2]*data$size)),
               log = TRUE))
}
## find the maximum likelihood estimate for the unknown parameters
optim(fn = llh, par = c(-1, 0.5), data = data)
## Compare to the ungrouped model in the course guide
glm_mod_ug <- glm(survived ~ size, family = "binomial", data = data)
coef(glm_mod_ug)
## Briefly comment on why, or why not, your parameter estimates are the same.
## Has the same model been fitted in each case?
## Plot your fitted models and compare!
```

