# Week 05 - Lecture 01

## Rough Agenda

### **4:05 pm - 5:55 pm** 
 
 + Modeller vs Designer (@sec-mods)
 + Multiple comparisons (@sec-multi)
 + Introducing Random Effects (@sec-randomeff)
    
### Before we begin **sort yourselves into groups** following @sec-groups


::: {.callout-tip}
## It's all about variation
 + Discuss the role of Sums of Squares (SS) in variance decomposition
 + Discuss the issues associated with multiple comparison techniques and the standard adjustment techniques employed to combat the issues
 + Apply and compare standard adjustment techniques (e.g., Tukey, Bonferroni and Fisher corrections) employed in multiple testing
 + Discuss and compare definitions of fixed and random effects

:::

## Modeller vs Designer {#sec-mods}

<p align="center">
<iframe src="slides/05-01_modller-vs-designer.pdf" height="400" width="700" title="Modeller-vs-Designer slides"></iframe>
</p>
<p align="center">
</p>


### Degrees of Freedom 

The **degree of freedom** is

 + the number of variables that are **free to vary** without breaking a constraint,
 + the number of **independent** observations or measurements that can be made in order to calculate some statistic.

:::{.callout-warning icon=false}
## <i class="fa fa-question-circle" aria-hidden="true"></i> <i class="fa fa-users" aria-hidden="true"></i> <i class="fa fa-pencil-square" aria-hidden="true"></i> Degrees of Freedom Quiz

In [StatChat](https://canvas.auckland.ac.nz/courses/141056/external_tools/38296) answer the questions as they are posed. 
:::


### Partitioning of Variation

::: {.callout-note collapse="true"}
## Variance Partition plot `R` function

```{r, eval = FALSE}
#' A function I wrote to plot the Sums of Squares
#' from ANOVA output showing the partition
#' @param x Either an object of class 'aov' or 'anova'
#' @param upset Logical, if TRUE then an upset plot rather than a
#' Venn diagram is drawn
require(venneuler)
require(stringr)
require(UpSetR)
vova <- function(x, upset = FALSE){
    if("aov" %in% class(x)) {
        ss = summary(x)[[1]]
    }else{
        if("anova" %in% class(x)){
            ss = x
        }else{
            stop("Unknown input: x mush be of class 'aov' or 'anova'.")
        }}
    n = nrow(ss)
    rnms = stringr::str_trim(rownames(ss)[1:(n-1)])
    nms = paste("SS(", rnms,")", sep = "")
    res = ss["Residuals", "Sum Sq"]
    pt = eval(paste(nms, "=", round(ss$"Sum Sq"[1:(n-1)], 2)))
    if(length( grep(":", rnms)) > 0) rnms =  stringr::str_replace(rnms, ":", "&")
    expr = setNames(round(ss$"Sum Sq"[1:(n-1)], 4), rnms)
    if(!upset){
        v = venneuler::venneuler(expr)
        plot(v)
        box()
        legend("bottomright", bty = "n", legend  = paste("SS(Residuals) =", round(res, 2)))
        legend("topleft", bty = "n", legend  = pt)
    }else{
        if(upset){
            UpSetR::upset(UpSetR::fromExpression(expr), sets.x.label = "SS(Main effects)",
                          mainbar.y.label = "SS(Interaction)", nsets = n)
        }
    }
}
```

:::

**In your groups** run through the `R` code below that fits a range of models to [data from Module 3 the courseguide](https://biosci738.github.io/BIOSCI738/module-3.html) and discuss/answer each question posed in the comments. Work out what each line of code does and replace the comments with your own more informative ones.

Be prepared to discuss your progress/thoughts with me as I wander around the class. NOTE: you will need the `R` function `vova()` from the drop down box above.


```{r, eval = FALSE}
## NOTE you will need the function vova() from the drop down above
url <- "https://raw.githubusercontent.com/STATS-UOA/databunker/master/data/factorial_expt.csv"
data <- readr::read_csv(url)
## example 1
lm(logAUC ~ Disease*Organ, data = data) |> aov() |> summary()
lm(logAUC ~ Disease*Organ, data = data) |> aov() |> vova()
## example 2
lm(logAUC ~ Organ*Disease, data = data) |> aov() |> summary()
lm(logAUC ~ Organ*Disease, data = data) |> aov() |> vova()
## example 3 
lm(logAUC ~ Organ + Disease, data = data) |> aov() |> summary()
lm(logAUC ~ Organ + Disease, data = data) |> aov() |> vova()
################
## unbalanced ##
################
unbal <- data[-c(1, 3),]
## example 1
lm(logAUC ~ Disease*Organ, data = unbal) |> aov() |> summary()
lm(logAUC ~ Disease*Organ, data = unbal) |> aov() |> vova()
## example 2
lm(logAUC ~ Organ*Disease, data = unbal) |> aov() |> summary()
lm(logAUC ~ Organ*Disease, data = unbal) |> aov() |> vova()
## example 3 
lm(logAUC ~ Organ + Disease, data = unbal) |> aov() |> summary()
lm(logAUC ~ Organ + Disease, data = unbal) |> aov() |> vova()
## type II
lm(logAUC ~ Organ*Disease, data = unbal) |> car::Anova(type = 2) 
lm(logAUC ~ Organ*Disease, data = unbal) |> car::Anova(type = 2) |>  vova()
## Three way ANOVA Example only.
## example 1
lm(logAUC ~ Disease*Organ*Animal, data = data) |> aov() |> summary()
lm(logAUC ~ Disease*Organ*Animal, data = data) |> aov() |> vova()
## example 2
lm(logAUC ~ Organ*Animal + Disease, data = data) |> aov() |> summary()
lm(logAUC ~ Organ*Animal + Disease, data = data) |> aov() |> vova()
## type II
lm(logAUC ~ Organ*Disease + Animal, data = unbal) |> car::Anova(type = 2) 
lm(logAUC ~ Organ*Disease + Animal, data = unbal) |> car::Anova(type = 2) |>  vova()
```




## Multiple Comparisons {#sec-multi}


In your groups discuss and answer the following points/questions, which relate the the snippet of code that follows.

 1. What is each line doing? 
 2. What is the relevance of 0.05  in the final line of the loop?
 3. What happens if you change this value?
 4. What happens if you change `n`, and why?
 5. What is this an example of? Why is this the case? 
 
Now, amend the code to ``test" the same thing using another statistical test of your choice.
 
```{r, eval = FALSE}
n <- 10000
t1err <- 0
for (i in 1:n){
    set.seed(1432 + i)
    x <- rnorm(100, 0, 1)
   if (((t.test(x, mu = 0))$p.value)<=0.05) (t1err = t1err + 1)
}
cat("Type I error rate in percentage is", (t1err/n)*100,"%", "\n")

```

:::{.callout-warning icon=false}
## <i class="fa fa-question-circle" aria-hidden="true"></i> <i class="fa fa-users" aria-hidden="true"></i> <i class="fa fa-pencil-square" aria-hidden="true"></i> Explain the phenomena above for an [r/ExplainLikeImFive](https://www.reddit.com/r/explainlikeimfive/?rdt=41742) audience

Open [StatChat](https://canvas.auckland.ac.nz/courses/141056/external_tools/38296) and **individually** write out your *[r/ExplainLikeImFive](https://www.reddit.com/r/explainlikeimfive/?rdt=41742)* (aka as if to a 5 year old child) reasoning for the phenomena illustrated by the code above.
:::


### Tukey vs. Bonferroni vs. Fisher

**Calculating (95%) Confidence Intervals**

Typically:

$$
\text{Upper bound} = \text{estimate} + (\text{scale factor} \times SE)
$$

$$
\text{Lower bound} = \text{estimate} - (\text{scale factor} \times SE)
$$

Now, imagine our quantity of interest is the *difference between means* (i.e., $\text{difference}$) :

$$
\text{Upper bound} = \text{difference} + (\text{scale factor} \times SED)
$$

$$
\text{Lower bound} = \text{difference} - (\text{scale factor} \times SED)
$$

OK, but how do we choose our $\text{scale factor}$? We have three^[actually we have many more but for the purposes of this course we have three, see @multicomp for discussion on some others] options:


|  | [Fisher (LSD)](https://en.wikipedia.org/wiki/Fisher%27s_least_significant_difference)^[A bunch of post hoc t-tests] | [Bonferroni Correction](https://en.wikipedia.org/wiki/Bonferroni_correction) | [Tukey (HSD)](https://en.wikipedia.org/wiki/Tukey%27s_range_test) |
|--|-------------|------------|-------------|
| $\text{scale factor}$ | $t_{\alpha = \frac{\alpha_c}{2}, \, df = N -m}$ | $t_{\alpha = \frac{\alpha_c}{2\cdot k}, \, df= N -m}$ | $q_{\alpha = 1- \alpha_c, \, m, \, df = N -m}$ |


where, 

- $N$ the number of observations
- $m$ the number of *treatment* groups  
- $SED$ standard error of the difference between means

   - $SED = \sqrt{MSE \left(\frac{1}{n_i} + \frac{1}{n_j}\right)}$, where $n_i, n_j$ are ample sizes of groups $i$ and $j$ respectively
  - $SED_{\text{equal group sizes}} = \sqrt{\dfrac{MSE}{n}}$
    
- $k = {m\choose2} = \frac{m(m-1)}{2}$ the number of pairwise comparisons being made 
- $\alpha_c$ the per comparison error rate (e.g., 0.05) 
- $df$ degrees of freedom   
- $t_{\alpha, df}$: Critical value from the t-distribution  
- $q_{\alpha, m, df}$: Critical value from the studentized range distribution (Tukey)


::: {.callout-note collapse="true"}
## Comparing critical values and CI coverage

```{r, echo = FALSE, message = FALSE}

library(dplyr)
library(tidyr)
library(ggplot2)

df <- 25
alpha <- 0.05
m_vals <- 2:10

results <- lapply(m_vals, function(m) {
  C <- m * (m - 1) / 2
  t_fisher <- qt(1 - alpha/2, df)
  t_bonf   <- qt(1 - (alpha/C)/2, df)
  t_tukey  <- qtukey(1 - alpha, m, df)/sqrt(2)
  pecr_fisher <- alpha
  pecr_bonf   <- alpha / C
  pecr_tukey  <- 1 - (1 - alpha)^(1/C)
  cov_fisher <- (1 - alpha)^C
  cov_bonf   <- 1 - alpha
  cov_tukey  <- 1 - alpha
  data.frame(
    m = m,
    comparisons = C,
    method = c("Fisher (t)", "Bonferroni (t)", "Tukey"),
    crit = c(t_fisher, t_bonf, t_tukey),
    coverage = c(cov_fisher, cov_bonf, cov_tukey),
    PECR = c(pecr_fisher, pecr_bonf, pecr_tukey),
    FWER = 1 - c(cov_fisher, cov_bonf, cov_tukey),
    power = 1 / c(t_fisher, t_bonf, t_tukey)
  )
}) |> bind_rows()

p1 <- ggplot(results, aes(m, crit, colour = method)) +
  geom_line(linewidth = 1.2, lty = "dotdash", alpha = 0.5) +
  geom_point(size = 2) +
  scale_colour_brewer(palette = "Set1") +
  labs( x = "Number of groups",y = "Critical Value",title = "", colour = "Method") +
  theme_minimal()
p2 <- ggplot(results, aes(m, PECR, colour = method)) +
  geom_line(linewidth = 1.2, lty = "dotdash", alpha = 0.5) +
  geom_point(size = 2) +
  scale_colour_brewer(palette = "Set1") +
  labs( x = "Number of groups",y = "PECR",
        title = "Error rate for one comparison", colour = "Method") +
  theme_minimal()
p3 <- ggplot(results, aes(m, coverage, colour = method)) +
  geom_line(linewidth = 1.2, lty = "dotdash", alpha = 0.5) +
  geom_point(size = 2) +
  scale_colour_brewer(palette = "Set1") +
  labs( x = "Number of groups",y = "Coverage",
        title = "Probability all intervals are correct", colour = "Method") +
  theme_minimal()

p4 <- ggplot(results, aes(m, FWER, colour = method)) +
  geom_line(linewidth = 1.2, lty = "dotdash", alpha = 0.5) +
  geom_point(size = 2) +
  scale_colour_brewer(palette = "Set1") +
  labs( x = "Number of groups",y = "FWER",
        title = "Error rate for any comparison", 
        colour = "Method") +
  theme_minimal()

p5 <- ggplot(results, aes(m, power, colour = method)) +
  geom_line(linewidth = 1.2, lty = "dotdash", alpha = 0.5) +
  geom_point(size = 2) +
  scale_colour_brewer(palette = "Set1") +
  labs( x = "Number of groups",y = "Power ",
        title = "Probability detect real differences", colour = "Method") +
  theme_minimal()

library(patchwork)

(p1 / (p2 | p3)) + plot_layout(guides = "collect") &
theme(legend.position = "bottom")

```

:::


**Code Along**

```{r, eval = FALSE}
library(tidyr)
library(ggplot2)
set.seed(1)
alpha <- 0.05
n <- 20 ## number of observations
nsim <- 500 ## number of simulations
m_vals <- 2:8 ## number of groups
## initialise empty vectors
det_fisher <- det_tukey <- det_bonf <- numeric(length(m_vals))
err_fisher <- err_tukey <- err_bonf <- numeric(length(m_vals))
## simulation loop
for(i in seq_along(m_vals)) {
  m <- m_vals[i]
  ## start "detect" counter at 0
  fisher_detect <- tukey_detect  <- bonf_detect   <- 0
  fisher_error <- tukey_error  <- bonf_error   <- 0
  for(sim in 1:nsim) {
    ## One TRUE difference
    means <- c(0.5, rep(0, m-1)) ## one 'true' difference
    y <- rnorm(n*m, mean = rep(means, each = n))
    ## check boxplot(y~rep(1:length(means), each = n))
    group <- factor(rep(1:m, each = n))
    ## Fisher (no correction - t-test)
    fisher <- pairwise.t.test(y, group, p.adjust.method = "none")
    if(any(fisher$p.value < alpha, na.rm = TRUE)) {
      fisher_detect <- fisher_detect + 1
    }
    ## Bonferroni
    bonf <- pairwise.t.test(y, group, p.adjust.method = "bonferroni")
    if(any(bonf$p.value < alpha, na.rm = TRUE)) {
      bonf_detect <- bonf_detect + 1
    }
    ## Tukey
    fit <- aov(y ~ group)
    if(any(TukeyHSD(fit)$group[, "p adj"] < alpha)) {
      tukey_detect <- tukey_detect + 1
    }
    ## No difference
    y <- rnorm(n*m, mean = 0)
    fisher <- pairwise.t.test(y, group, p.adjust.method = "none")
    bonf   <- pairwise.t.test(y, group, p.adjust.method = "bonferroni")
    tukey  <- TukeyHSD(aov(y ~ group))
    fisher_error <- fisher_error + any(fisher$p.value < alpha, na.rm = TRUE)
    bonf_error   <- bonf_error   + any(bonf$p.value   < alpha, na.rm = TRUE)
    tukey_error  <- tukey_error  + any(tukey$group[, "p adj"] < alpha)
  }
  det_fisher[i] <- fisher_detect/nsim
  det_bonf[i]   <- bonf_detect/nsim
  det_tukey[i]  <- tukey_detect/nsim
  err_fisher[i] <- fisher_error/nsim
  err_bonf[i]   <- bonf_error/nsim
  err_tukey[i]  <- tukey_error/nsim
}
power_df <- data.frame(m = m_vals, Fisher = det_fisher, Bonferroni = det_bonf, Tukey = det_tukey)
error_df <- data.frame(m = m_vals, Fisher = err_fisher, Bonferroni = err_bonf, Tukey = err_tukey)
power_df |>
  pivot_longer(-m, names_to = "method", values_to = "detect") |>
  ggplot(aes(m, detect, colour = method)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_colour_brewer(palette = "Set1") +
  labs( x = "Number of groups",y = "Proportion of times a difference detected", colour = "Method") +
  theme_minimal() 
error_df |>
  pivot_longer(-m, names_to = "method", values_to = "detect") |>
  ggplot(aes(m, detect, colour = method)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_colour_brewer(palette = "Set1") +
  labs( x = "Number of groups",y = "Family-wise error rate", colour = "Method") +
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  theme_minimal() 
```

#### Group Activity

```{r,eval = FALSE}
data <- data.frame(group = factor(rep(c("A","B","C","D"), each = n)),
                   y = c(12.7419168942933, 8.87060365720782, 
10.7262568226747, 11.2657252099221, 10.808536646282, 9.78775096781703, 
13.0230439948779, 9.8106819231738, 14.0368474277541, 9.87457180189516, 
12.609739308447, 14.5732907854022, 8.42227859777532, 10.6424224663653, 
10.9333573272127, 12.4719007961401, 10.6314941571679, 5.88708915819045, 
6.31906614284896, 13.8402266914604, 10.586722811843, 7.63738313204, 
10.8561652884808, 13.6293493983452, 16.7903869225299, 12.1390617367876, 
12.4854612344621, 9.47367382961044, 13.9201947096625, 11.7200102480798, 
13.9109002464824, 14.4096746744576, 15.0702070439398, 11.7821472491856, 
14.0099102465959, 9.56598264185332, 8.93108198324101, 8.79818481164696, 
5.67158470010674, 10.5722452137845, 10.9119972004005, 9.77788540290267, 
12.016326471399, 9.04659034584685, 7.76343791116141, 11.3656360517774, 
8.87721364762666, 13.3882025234425))
```


```{r, eval = FALSE, eval = FALSE}

lm <- lm(y ~ group, data = data)
pm <- predictmeans::predictmeans(lm, modelterm = "group", pairwise = TRUE, plot = FALSE, level = 0.05/choose(4,2), adj = "bonferroni")
comparisons(pm)
```

1. Using the data above perform the pairwise comparisons of means using Fisher’s LSD with $\alpha = 0.05$ level of significance. 

2. Summarise and present these results in a table with the following column names **Comparison**, **Calculated difference** (in means), **SED** (standard error of the difference), **LSDs**, **Lower 95\% CI** &  **Upper 95\% CI** (lower & upper 95\% confidence interval), **statistic**, **p-value**.

3. Calculate the scale factor (i.e., critical value of the assumed distribution) using 1) Bonferroni's correction method and 2) Tukey's HSD method for any pairwise comparison 95\% CI from above. Discuss what effect the multipliers might have on inference by constructing and comparing the respective pairwise comparisons tables (e.g., as in 2).

NOTE: you may find [this section of the course guide](https://biosci738.github.io/BIOSCI738/multiple-comparisons.html#adjustments-for-multiple-testing) useful for this activity; in particular [this function I wrote might be of use](https://gist.github.com/cmjt/72f3941533a6bdad0701928cc2924b90), which can be loaded into `R` using `url <- "https://gist.github.com/cmjt/72f3941533a6bdad0701928cc2924b90"` and 
`devtools::source_gist(url, quiet = TRUE)`.



```{r, eval = FALSE, echo = FALSE}
url <- "https://gist.github.com/cmjt/72f3941533a6bdad0701928cc2924b90"
devtools::source_gist(url, quiet = TRUE)
lm <- lm(y ~ group, data = data)
pm <- predictmeans::predictmeans(lm, modelterm = "group", pairwise = TRUE, plot = FALSE)
comparisons(pm)
pm <- predictmeans::predictmeans(lm, modelterm = "group", pairwise = TRUE, plot = FALSE, level = 0.05/choose(4,2), adj = "bonferroni")
comparisons(pm)
m <- length(levels(data$group))
C <- m * (m - 1) / 2
alpha <- 0.05
t_fisher <- qt(1 - alpha/2, df)
t_bonf <- qt(1 - (alpha/C)/2, df)
t_tukey <- qtukey(1 - alpha, m, df)/sqrt(2)
```

## Introducing random effcets {#sec-randomeff}

**Slides** - *code along*

```{r, eval = FALSE}
base_url <- "https://raw.githubusercontent.com/STATS-UOA/databunker/master/data/"
rcbd <- readr::read_csv(paste(base_url, "rcbd.csv", sep = ""))
rcbd$Run <- as.factor(rcbd$Run)
lm <- lm(logAUC8 ~ Run + Surgery, data = rcbd)
lm |> summary()
anova(lm)
lmer4_mod <- lme4::lmer(logAUC8 ~ Surgery + (1|Run), data = rcbd)
coefficients(lmer4_mod)
lmer4_mod |> summary()
```




<p align="center">
<iframe src="slides/05-04_fixed-vs-random.pdf" height="400" width="700" title="Fixed vs Random effects slides"></iframe>
</p>
<p align="center">
</p>

Below is a list collated by @gelman2005analysis (p. 20) who discuss fixed and random effects in the context of ANOVA-type analysis (Note that some use terminology we're yet to cover!).

> 1. Fixed effects are constant across individuals, and random effects vary. For example, in a growth study, a model with random intercepts $\alpha_i$ and fixed slope $\beta$ corresponds to parallel lines for different individuals $i$, or the model $y_{it} = \alpha_i + \beta t$. Kreft and de Leeuw [(1998), page 12] thus distinguish between fixed and random coefficients.
2. Effects are fixed if they are interesting in themselves or random if there is interest in the underlying population. Searle, Casella, and McCulloch [(1992), Section 1.4] explore this distinction in depth.
3. When a sample exhausts the population, the corresponding variable is *fixed*; when the sample is a small (i.e., negligible) part of the population, the corresponding variable is *random* [Green and Tukey (1960)].
4. If an effect is assumed to be a realized value of a random variable, it is called a random effect" [LaMotte (1983)].
5. Fixed effects are estimated using least squares (or, more generally, maximum likelihood), and random effects are estimated with shrinkage ["linear unbiased prediction" in the terminology of Robinson (1991)]. This definition is standard in the multilevel modeling literature [see, e.g., Snijders and Bosker (1999), Section 4.2] and in econometrics. In the Bayesian framework, this definition implies that fixed effects $\beta_j^{(m)}$ are estimated conditional on $\sigma_m = \infty$, and random effects $\beta_j^{(m)}$ are estimated conditional on $\sigma_m$ from the posterior distribution.


The author of @gelman2005analysis moves on to the following conclusion (see p. 21).

> We prefer to sidestep the overloaded terms "fixed" and "random" with a cleaner distinction by simply renaming the terms in definition 1 above. We define effects (or coefficients) in a multilevel model as **constant** if they are identical for all groups in a population and **varying** if they are allowed to differ from group to group. For example, the model  $y_{ij} = \alpha + \beta x_{ij}$ (of units $i$ in groups $j$) has a constant slope and varying intercepts, and  $y_{ij} = \alpha_{j} + \beta_j x_{ij}$ has varying slopes and intercepts. In this terminology (which we would apply at any level of the hierarchy in a multilevel model), varying effects occur in batches, whether or not the effects are interesting in themselves (definition 2), and whether or not they are a sample from a larger set (definition 3). Definitions 4 and 5 do not arise for us since we estimate all batches of effects hierarchically, with the variance components $\sigma_m$ estimated from data.



**As a group**

1. Translate each definition in the list above to lay-person language (i.e., non-technical).
2. Make up a data example and allocate your fixed and random effects according to each definition.
3. Discuss the similarities and dissimilarities between the definitions.
4. Conclude with which definition *best* aligns with your current understanding of the distinction between fixed and random effects?

NOTE: You may find [this blog](https://dynamicecology.wordpress.com/2015/11/04/is-it-a-fixed-or-random-effect/) offers some extra insights here^[Note that I do like the writing style of the blog but don't totally agree with the author's conclusions around repeated measurements; if you want to know why, ask me!]

:::{.callout-warning icon=false}
## <i class="fa fa-question-circle" aria-hidden="true"></i> <i class="fa fa-users" aria-hidden="true"></i> <i class="fa fa-pencil-square" aria-hidden="true"></i> Explain fixed vs. random effects to an [r/ExplainLikeImFive](https://www.reddit.com/r/explainlikeimfive/?rdt=41742) audience

Open [StatChat](https://canvas.auckland.ac.nz/courses/141056/external_tools/38296) and **individually** write out your *[r/ExplainLikeImFive](https://www.reddit.com/r/explainlikeimfive/?rdt=41742)* (aka as if to a 5 year old child) distinction!
:::
