# Week 05 - Lecture 02

## Rough Agenda

### **2:05 pm - 3:55 pm** 

+ Interactive Oral assessments (@sec-ioa)
+ Experimental Design (@sec-edterms)

 
### Before we begin **sort yourselves into groups** following @sec-groups

::: {.callout-tip}
## Experiments and communication
 + Communicate data structure and statistical hypotheses
 + Create an experimental design 'recipe' using `R`
:::



## Interactive Oral assessments (IOAs) {#sec-ioa}

*In-class content*

::: {.callout-note}
### What is an IOA

> IOAs are novel, low-stakes authentic assessments, designed to promote academic integrity and prepare students for the workplace. Communication as a key skill in the UoA's graduate profile and is integral for students' skill development and employability. An IOA is authentic because it is based on a real-world scenario and promotes student engagement and facilitates higher-order thinking. It also preserves academic integrity through its unscripted conversation prompts based on student responses.

:::

:::{.callout-warning icon=false}
## <i class="fa fa-question-circle" aria-hidden="true"></i> <i class="fa fa-users" aria-hidden="true"></i> <i class="fa fa-pencil-square" aria-hidden="true"></i> Grade a mock IOA


**Choose one of the mock IOAs on [this CANVAS page](https://canvas.auckland.ac.nz/courses/141056/pages/ioas-in-biosci738)**

 1. Watch your chosen mock IOA and discuss as a group how you would grade Rishika according to the given rubric, and
 2. *Grade* Rishika according to [the rubric for your Interactive Oral Assessment - Project Discussion](https://canvas.auckland.ac.nz/courses/141056/assignments/490477)
 3. Open [StatChat](https://canvas.auckland.ac.nz/courses/141056/external_tools/38296) and **individually** write out your *grades* for each rubric category.
 
Feel free to ask any IOA related questions to either Adam or I as we wander around the class.

:::

#### Group Activity

**Discuss the [Biota of University of Auckland City Campus iNaturlist data](https://www.inaturalist.org/projects/biota-of-university-of-auckland-city-campus)**

If you've not already Sign Up and download any subset of [these data](ttps://www.inaturalist.org/projects/biota-of-university-of-auckland-city-campus).


In pairs within your group, take on the following roles in turn **Consultant** and **Client**.

As the **Client** answer the following questions about the iNaturalist data for you [Significance Article + Code](https://canvas.auckland.ac.nz/courses/141056/assignments/490408).

 + What variables are included in the dataset?
 + What are the potential issues with these citizen science data?
 + What questions are you trying to answer with these data? 
 + What other data could you use to supplement any analysis (e.g., topography, weather)?

As the **Consultant**, your role is to question the **Client** and ask for specific details/clarifications along the way. After the Q&A session, you should be able to surmise your **Client's** project in your own words.




::: {.callout-important}
## Don't forget to sign up for your [Interactive Oral Assessment - Project Discussion](https://canvas.auckland.ac.nz/courses/141056/assignments/490477) time slot via BIOSCI738's CANVAS calander.
:::


## Experimental Design {#sec-edterms}

### The Grammar of Experimental Designs^[The title stolen shamelessly from [Emi Tanaka's ebook](https://emitanaka.org/edibble-book/)] 

This section briefly quizzes you on the terms and concepts discussed in [this section of the course guide](https://biosci738.github.io/BIOSCI738/introduction-to-the-design-and-analysis-of-experiments.html) and [Emi Tanaka's `edibble` ebook](https://emitanaka.org/edibble-book/). Refer tho these sources for more details definitions. 


**Experimental design quiz**

<style>
.match-container {
  display: grid;
  grid-template-columns: 1fr 1fr;
  gap: 20px;
  margin-top: 20px;
}

.match-item, .drop-zone {
  padding: 10px;
  border-radius: 6px;
  margin-bottom: 10px;
}

.match-item {
  background: #e8f4f8;
  border: 1px solid #ccc;
  cursor: grab;
}

.drop-zone {
  border: 2px dashed #bbb;
  min-height: 40px;
  background: #fafafa;
}

.correct {
  background: #d4edda;
}

.incorrect {
  background: #f8d7da;
}
</style>


<div class="match-container">

<!-- draggable answers -->
<div>
<div class="match-item" draggable="true" data-match="obj">Direct the experiment by stating exactly what you aim to measure or test.</div>

<div class="match-item" draggable="true" data-match="resp">The variable you measure to assess the effect of treatments.</div>

<div class="match-item" draggable="true" data-match="factor">A variable you manipulate to see its effect on the response.</div>

<div class="match-item" draggable="true" data-match="list">Helps organize and identify variables you can control.</div>

<div class="match-item" draggable="true" data-match="material">The subjects or units on which treatments are applied.</div>

<div class="match-item" draggable="true" data-match="shared">When subjects/units share conditions, measurements may not be independent</div>
</div>

<!-- drop zones (questions) -->
<div>

<p><strong>Why are specific objectives important?</strong></p>
<div class="drop-zone" data-accept="obj"></div>

<p><strong>What is the response variable?</strong></p>
<div class="drop-zone" data-accept="resp"></div>

<p><strong>What is an experimental factor?</strong></p>
<div class="drop-zone" data-accept="factor"></div>

<p><strong>Why list the experimental factors?</strong></p>
<div class="drop-zone" data-accept="list"></div>

<p><strong>What is the experimental material?</strong></p>
<div class="drop-zone" data-accept="material"></div>

<p><strong>Why is a 'shared environment' an issue?</strong></p>
<div class="drop-zone" data-accept="shared"></div>

</div>
</div>

<button onclick="checkAnswers()">Check Answers</button>

<script>
const draggables = document.querySelectorAll('.match-item');
const dropzones = document.querySelectorAll('.drop-zone');

draggables.forEach(item => {
  item.addEventListener('dragstart', e => {
    e.dataTransfer.setData('text/plain', item.dataset.match);
    window.dragged = item;
  });
});

dropzones.forEach(zone => {
  zone.addEventListener('dragover', e => e.preventDefault());
  zone.addEventListener('drop', e => {
    e.preventDefault();
    zone.innerHTML = "";
    zone.appendChild(window.dragged);
  });
});

function checkAnswers() {
  dropzones.forEach(zone => {
    const item = zone.querySelector('.match-item');
    if (!item) return;

    if (item.dataset.match === zone.dataset.accept) {
      zone.classList.add('correct');
      zone.classList.remove('incorrect');
    } else {
      zone.classList.add('incorrect');
      zone.classList.remove('correct');
    }
  });
}
</script>


::: {.callout-note}
## Out of lecture activity!
Read and critique this blog! [Pseudoreplication: choose your data wisely](https://www.statisticsdonewrong.com/pseudoreplication.html)
:::

### Design an alien plant experiment

**Scenario**

Imagine that you are a plant biologist testing the effect of four fertilizer types on the growth of an alien plant species. Below are the design features of your setup!

 + **60** seedling alien plants (*the species name is yet to be decided*)
 + **4** fertilizer types: *martian_soil*, *uranium*, *sulphur*, and *solar_plasma*
 + **3** different greenhouses (*20* spaces in each): 


**In your groups** discuss

   + what you are measuring,
   + the experimental units, and
   + the sources of variation.

Then choose one or more of the following `R` packages (`edibble`, `agricolae` or `AlgDesign`) to structure/design the above experimental scenario (i.e., figure out how to allocated treatments to units etc.).
 
NOTE: you may find [this application useful](https://statbiscuit.shinyapps.io/doe_form/) to structure your thoughts.


Now, once you've shown your design to Charlotte you'll get a keyword to access some data below!

```{r, echo = FALSE}
simulate_alien <- function(seed, fertilizer_effects) {
  
  set.seed(seed)
  
  fertilizers <- c("martian_soil", "uranium", "sulphur", "solar_plasma")
  n_greenhouses <- 3
  plants_per_fertilizer <- 5
  
  greenhouse_sd <- 0.3   
  plant_sd <- 0.5        
  
  data <- expand.grid(
    fertilizer = fertilizers,
    growth_chamber = paste0("growth_chamber_", 1:n_greenhouses),
    plant_id = 1:plants_per_fertilizer
  )
  
  greenhouse_effect <- rnorm(n_greenhouses, 0, greenhouse_sd)
  names(greenhouse_effect) <- paste0("growth_chamber_", 1:n_greenhouses)
  
  data$height <- mapply(function(fert, gh) {
    mean_height <- fertilizer_effects[fert]
    mean_height + greenhouse_effect[gh] + rnorm(1, 0, plant_sd)
  }, data$fertilizer, data$growth_chamber)
  
  data
}

data_triffids <- simulate_alien(
  101,
  c(
    martian_soil = 9,
    uranium = 10.3,
    sulphur = 11.5,
    solar_plasma = 10.2
  )
)

data_nebula <- simulate_alien(
  202,
  c(
    martian_soil = 9.5,
    uranium = 10.8,
    sulphur = 10.9,
    solar_plasma = 9.9
  )
)

data_xenophyte <- simulate_alien(
  303,
  c(
    martian_soil = 10.0,
    uranium = 10.2,
    sulphur = 10.4,
    solar_plasma = 10.1
  )
)

data_cryomoss <- simulate_alien(
  404,
  c(
    martian_soil = 9.7,
    uranium = 11.2,
    sulphur = 10.5,
    solar_plasma = 10.3
  )
)

```


<input type="text" id="secretAlien" placeholder="Enter keyword here">
<button onclick="checkKeywordAlien()">👽🌵 Get the data 🌵👽</button>

<div id="Triffids" style="display:none; margin-top:10px;">

```{r, echo = FALSE}
datapasta::df_paste(data_triffids)
```
</div>

<div id="Nebula" style="display:none; margin-top:10px;">
```{r, echo = FALSE}
datapasta::df_paste(data_nebula)
```
</div>

<div id="Xenophyte" style="display:none; margin-top:10px;">
```{r, echo = FALSE}
datapasta::df_paste(data_xenophyte)
```
</div>

<div id="Crymoss" style="display:none; margin-top:10px;">
```{r, echo = FALSE}
datapasta::df_paste(data_cryomoss)
```
</div>

<script>
function checkKeywordAlien() {
  var input = document.getElementById("secretAlien").value;
  
  document.getElementById("Triffids").style.display = "none";
  document.getElementById("Nebula").style.display = "none";
  document.getElementById("Xenophyte").style.display = "none";
  document.getElementById("Crymoss").style.display = "none";
  
  if(document.getElementById(input)) {
    document.getElementById(input).style.display = "block";
  } else {
    alert("Incorrect keyword!");
  }
}
</script>

**As a group** model the data *unlocked* above and

  + create a plot comparing the treatment effects (*if any*),
  + create a plot that shows the partitioning of variance for you chosen model.
  
Your **reporter** (remember @sec-groups) will then give a presentation to the class talking us through the plots! 






