# Activity solutions {#sec-solutions}

::: {.callout-note}
## Suggested `R` code solutions

This section will be update with my suggested solutions to the class wide activities after their conclusion. I **strongly** recommend that you review these outside of class if only to familiarise yourself with a different approach.
:::


## Suggested solution to @sec-gorilla

```{r, eval = FALSE}
require(tidyverse)
## read in data
data <- readr::read_csv("https://raw.githubusercontent.com/STATS-UOA/databunker/master/data/BMI.csv")
data
## Mean BMI
data %>%
    summarise(mean(bmi)) %>%
    round()
## Max BMI males
data %>%
    filter(gender == "male") %>%
    filter(bmi == max(bmi))
## t-test BMI males vs females
t.test(bmi ~ gender, data = data)
lm(bmi ~ gender, data = data) |> summary()
## t-test steps males vs females
t.test(steps ~ gender, data = data) 
lm(steps ~ gender, data = data) |> summary()
## plotting. ALWAYS PLOT YOUR DATA!
ggplot(data, aes(x = steps, y = bmi, col = gender)) +
    geom_point()
```

## Suggested solution to @sec-draw

```{r, eval = FALSE}
## data source: https://docs.google.com/spreadsheets/d/1YSJ4ypkYLq6j1mIBJCgUHhHjJZQ0Rkfe1qW2WC5HLiw/edit#gid=748627588
library(tidyverse)
data <- read_csv("https://raw.githubusercontent.com/STATS-UOA/databunker/master/data/marvel_movies.csv") %>%
	mutate(critics_score = as.numeric(str_replace_all(`critics % score`, "%", ""))) %>%
	mutate(budget_recovered = as.numeric(str_replace_all(`% budget recovered`, "%", "")))
##### ##### ##### ###
##### Model 1 #######
##### ##### ##### ###
data %>%
	lm(budget_recovered ~ critics_score, data = .) |>
	summary()
## Ans
data %>%
	ggplot(., aes(y = budget_recovered, x = critics_score)) +
	geom_smooth(method = "lm", se = FALSE) + ylab(" Estimated budget_recovered")
## w. data
data %>%
	ggplot(., aes(y = budget_recovered, x = critics_score)) +
	geom_smooth(method = "lm", se = FALSE) + geom_point()
##### ##### ##### ###
##### Model 2 #######
##### ##### ##### ###
names(table(data$category))
mod <- data %>%
	lm(budget_recovered ~  category, data = .) 
mod |> summary()
## Ans
data$pred_vals <- predict(mod)
## plot
data %>%
	ggplot(aes(y = pred_vals, x = category)) +
    geom_point(size = 15, pch = "-") + ylab(" Estimated budget_recovered")

##### ##### ##### ###
##### Model 3 #######
##### ##### ##### ###
mod <- data %>%
	lm(budget_recovered ~ critics_score + category, data = .) 
mod |>	summary()
## Ans
data$pred_vals <- predict(mod)
## plot
data %>%
	ggplot(aes(y = pred_vals, x = critics_score, color = category)) +
	geom_line() + ylab(" Estimated budget_recovered")

##### ##### ##### ###
##### Model 4 #######
##### ##### ##### ###
data <- data %>%
	rename(., "worldwide" = `worldwide gross ($m)`, "domestic" = `domestic gross ($m)`,
				 "international" = `international gross ($m)`)
mod <- lm(worldwide ~ 0 + domestic + international, data = data) 
mod |> summary()
## Ans
## plot
require(rsm) ## install if not available
persp(mod, form =  ~ 0 + domestic + international)
image(mod, form =  ~ 0 + domestic + international)
##### ##### ##### ###
##### Model 5 #######
##### ##### ##### ###
mod <- lm(worldwide ~ budget*critics_score, data = data) 
mod |>	summary()
## Ans
## plot
persp(mod, form =  ~ budget*critics_score)
image(mod, form =  ~ budget*critics_score)
```


## Some suggested in-class metrics for @sec-taylor

**Number of time until the first repeat** (in this case *Blurple*)

```{r, eval = FALSE}
final_leg_outfits <- c("Cotton candy", "Blurple", "Grapefruit", "Popsicle", "Sunset orange", 
"Blurple", "Cotton candy", "Popsicle", "Grapefruit", "Popsicle", 
"Blurple", "Grapefruit", "Cotton candy", "Popsicle", "Blurple", 
"Blurple", "Blurple", "Blurple")
## times until first repeat
## in this case Blurple, with 5 outfits before repeat
obvs <- which(duplicated(final_leg_outfits))[1] - 1
randomisation <- function(data, nreps = 1000, seed = 1984){
  sampling_dist <- numeric(nreps)
  set.seed(seed) 
  for (i in 1:nreps) {
   sampling_dist[i] <- (which(duplicated(sample(data)))[1] - 1)
  }
return(sampling_dist)
}
sampling_distribution <- randomisation(final_leg_outfits)
hist(sampling_distribution)
abline(v = obvs, lwd = 2)
```

**Number of times we have *Sunset Orange* thrice in a row**

```{r, eval = FALSE}
europe_leg_outfits <- c("Flamingo pink", "Ocean blue", "Sunset orange", "Ocean blue", 
"Flamingo pink", "Sunset orange", "Ocean blue", "Sunset orange", 
"Flamingo pink", "Sunset orange", "Ocean blue", "Flamingo pink", 
"Sunset orange", "Ocean blue", "Sunset orange", "Flamingo pink", 
"Flamingo pink", "Ocean blue", "Sunset orange", "Sunset orange", 
"Sunset orange", "Ocean blue", "Flamingo pink", "Ocean blue", 
"Flamingo pink", "Sunset orange", "Sunset orange", "Ocean blue", 
"Flamingo pink", "Sunset orange", "Flamingo pink", "Sunset orange", 
"Ocean blue", "Sunset orange", "Ocean blue", "Flamingo pink", 
"Sunset orange", "Ocean blue", "Flamingo pink", "Sunset orange", 
"Ocean blue", "Sunset orange", "Flamingo pink", "Ocean blue", 
"Flamingo pink", "Sunset orange", "Flamingo pink", "Sunset orange")
## Sunset orange only appears exactly thrice in a row once
run_lengths <- rle(europe_leg_outfits) ## inbuilt R function for run lengths
obvs <- sum(run_lengths$values == "Sunset orange" & run_lengths$lengths == 3)
randomisation <- function(data, nreps = 1000, seed = 1984){
  sampling_dist <- numeric(nreps)
  set.seed(seed) 
  for (i in 1:nreps) {
    x <- sample(data) |> rle()
   sampling_dist[i] <- sum(x$values == "Sunset orange" & x$lengths == 3)
  }
return(sampling_dist)
}
sampling_distribution <- randomisation(europe_leg_outfits)
hist(sampling_distribution)
abline(v = obvs, lwd = 2)
```

