# Activity solutions {#sec-solutions}

::: {.callout-note}
## Suggested `R` code solutions

This section will be update with my suggested solutions to the class wide activities after their conclusion. I **strongly** recommend that you review these outside of class if only to familiarise yourself with a different approach.
:::


## Suggested solution to @sec-gorilla

```{r, eval = FALSE}
require(tidyverse)
## read in data
data <- readr::read_csv("https://raw.githubusercontent.com/STATS-UOA/databunker/master/data/BMI.csv")
data
## Mean BMI
data %>%
    summarise(mean(bmi)) %>%
    round()
## Max BMI males
data %>%
    filter(gender == "male") %>%
    filter(bmi == max(bmi))
## t-test BMI males vs females
t.test(bmi ~ gender, data = data)
## t-test steps males vs females
t.test(steps ~ gender, data = data) 

## plotting. ALWAYS PLOT YOUR DATA!
ggplot(data, aes(x = steps, y = bmi, col = gender)) +
    geom_point()
```

## Suggested solution to @sec-honesty

```{r, eval = FALSE}
library(tidyverse)
data <- read_csv("retracted-honesty-study_week-02.csv")
## set up descriptive treatments
data <- data %>% mutate (treatment = ifelse(Cond==0, "No signature",
                                     ifelse(Cond==1, "Sign-Top",
                                     ifelse(Cond==2, "Sign-Bottom" , NA))))
## remove control condition (here we're only interested in comparing sign-top and sign-bottom)
data_sub <- data %>% filter(., Cond != 0)
## MATCH PAPER
## "lowered the share of people over-reporting their math puzzle performance from 79% to 37%"
## with "chi sq test p(0.0013)"
prop.table(table(data_sub$CheatedOnMatrixTax, data_sub$treatment), margin = 2)
table(data_sub$CheatedOnMatrixTax, data_sub$treatment) |> chisq.test()
## "lowered the average amount of over-reporting from 3.94 puzzles to 0.77 puzzles (p < .00001)"
t.test(OverReport ~ treatment, data = data_sub)
## "nearly halved the average amount of claimed commuting expenses, from $9.62 to $5.27 (p = .0014)"
t.test(SumDeduction ~ treatment, data = data_sub)

## DATA DETECTIVING
## lm
mod <- lm(SumDeduction ~ treatment, data = data_sub)
mod |> summary() ## matches t-test above (is the same thing)
## plot makes sense
ggplot(data_sub, aes(x = treatment, y = SumDeduction)) +
    geom_violin() + geom_jitter()
## mapping flagged observations to colour
ggplot(data_sub, aes(x = treatment, y = SumDeduction)) +
    geom_violin() + geom_jitter(aes(col = as.factor(flag)))
## residuals
data_sub$residuals <- residuals(mod)
ggplot(data_sub, aes(x = treatment, y = residuals)) +
    geom_point(aes(col = as.factor(flag))) +
    geom_hline(yintercept = 0, col = "grey", linewidth = 2, linetype = 2)
## now removing flagged observations
## lm
data_sub %>%
    filter(., flag == 0) %>%
    t.test(SumDeduction ~ treatment, data = .)

## see https://datacolada.org/109
## and/or https://www.theguardian.com/education/2023/jun/25/harvard-professor-data-fraud
## for the story


```